---
layout: remark
title: Introduction and Basics of R
subtitle: Data Analysis with R and Python
author: Deepayan Sarkar
mathjax: true
---

<!-- 

# Topics to be Covered

- Installing and running R and RStudio, using the command line console

- R as an advanced calculator with graphics and simulation capabilities

- Vectors (numeric, character, logical, factors) and indexing

- Lists and Data frames

- Operators and Functions, Attributes

- The Help System (and S3 methods)

-->


# Review: Software for Statistics

- Computation is an essential part of modern statistics

    - Handling large datasets 

    - Visualization

--
    - Simulation

    - Iterative methods 


--

- Many options, but we will focus on R and Python

	- Available as [Free](https://en.wikipedia.org/wiki/Free_software_movement) / 
      [Open Source](https://en.wikipedia.org/wiki/The_Open_Source_Definition) Software

	- Easy to try out on your own

	- Contains all essential data analysis tools
	
	- Active community
	
	

```{r opts, echo = FALSE, results = "hide", warning = FALSE, message = FALSE}
library(knitr)
library(kableExtra)
opts_chunk$set(cache = TRUE, cache.path='~/knitr-cache/darp-intro/', autodep = TRUE,
               comment = "", warning = TRUE, message = TRUE,
               fig.width = 12, fig.height = 5.5,
               dev = "svglite", dev.args = list(pointsize = 15, bg = "transparent"),
               knitr.table.format = "html",
               fig.path='figures/1-intro-')
opts_chunk$set(engine.path = list(python = '/usr/local/bin/python3'))
options(warnPartialMatchDollar = FALSE, width = 90) # 110
```


<div>
$$
\newcommand{\sub}{_}
$$
</div>


---

# Interacting with R

* R is most commonly used as a [REPL](https://en.wikipedia.org/wiki/Read-eval-print_loop) (Read-Eval-Print-Loop)

	* When it is started, R Waits for user input

	* User inputs expression

	* R evaluates and prints result

	* Waits for more input

--

* Python supports exactly the same model 

--

* Several _interfaces_ are available to help in this process

* Recommended options (representing slightly different approaches)

	* RStudio
	
	* Jupyter

---

# Running R

* On starting up, R shows you something like this:

```
R version 4.2.2 Patched (2022-11-17 r83375) -- "Innocent and Trusting"
Copyright (C) 2022 The R Foundation for Statistical Computing
Platform: x86_64-pc-linux-gnu (64-bit)

R is free software and comes with ABSOLUTELY NO WARRANTY.
You are welcome to redistribute it under certain conditions.
Type 'license()' or 'licence()' for distribution details.

R is a collaborative project with many contributors.
Type 'contributors()' for more information and
'citation()' on how to cite R or R packages in publications.

Type 'demo()' for some demos, 'help()' for on-line help, or
'help.start()' for an HTML browser interface to help.
Type 'q()' to quit R.

>
```

* The `>` represents a _prompt_ indicating that R is waiting for input

---

# Running Python

* Python has a similar but much shorter start-up message:

```
Python 3.9.12 (main, Apr  5 2022, 01:53:17) 
[Clang 12.0.0 ] :: Anaconda, Inc. on darwin
Type "help", "copyright", "credits" or "license" for more information.
>>> 
```

* The `>>>` similarly represents the _prompt_ indicating that Python is waiting


---

# The R REPL essentially works like a calculator

```{r}
34 * 23 + 10
27 + 1 / 7
2^10
exp(2)
```

---

# The Python REPL is very similar

```{python}
34 * 23 + 10
27 + 1 / 7
```

---

# But not exactly the same

```{python}
2^10
```

---

# But not exactly the same


```{python}
2^10
```

```{python}
2 ** 10
```

---

# But not exactly the same

```python
exp(2)
```

```
Traceback (most recent call last):
  File "<stdin>", line 1, in <module>
NameError: name 'exp' is not defined
```

---

# NumPy for scientific computing in Python

* Python with NumPy

```{python}
from numpy import *
exp(2)
sin(pi / 6)
```

* Compare with R

```{r}
exp(2)
sin(pi / 6)
```

---

# Why learn both R and Python?

* Both are just tools... skills are more important

--

* Complementary strengths

--

* Easy interoperability

---

# What about error messages?

```r
34 +* 23
```

```
Error: unexpected '*' in "34 +*"
```

```r
27 // 7
```

```
Error: unexpected '/' in "27 //"
```


```{r,error = TRUE}
arcsin(1/2)
```

---

class: center middle

# Essential features of R

---

layout: true

# R supports mathematical functions

---

```{r}
sqrt(5 * 125)
log(120)
factorial(10)
log(factorial(10))
```

---

```{r}
choose(15, 5)
factorial(15) / (factorial(10) * factorial(5))
```

--

```{r}
choose(1500, 2)
factorial(1500) / (factorial(1498) * factorial(2))
```

---

layout: false

# R supports variables

```{r}
x <- 2
y <- 10
x^y
y^x
factorial(y)
log(factorial(y), base = x)
```

---

# R can compute on vectors

```{r}
N <- 15
x <- seq(0, N)
N
x
1:N
choose(N, x)
```

---

# Example: Evaluating Binomial probabilities

- Binomial distribution: 

$$
P(X = x) = { n \choose x } p^x (1-p)^{n-x}
$$

--

```{r}
N <- 15
x <- seq(0, N)
p <- 0.25
choose(N, x) * p^x * (1-p)^(N-x)
```

---

# _Summary functions_ that work on vectors

```{r}
p.x <- dbinom(x, size = N, prob = p)
p.x
x * p.x
sum(x * p.x) / sum(p.x)
N * p
```


---

# Example: Factorials revisited

```{r}
choose(1500, 2)
log(choose(1500, 2))
factorial(1500) / (factorial(1498) * factorial(2)) # fails
```

--

```{r}
factorial(1500)
```

---

layout: true

# Factorials using vectorized arithmetic

---

```{r}
factorial(10)
prod(seq(1, 10))
```

--

```{r}
log(prod(seq(1, 10)))
sum(log(1:10))
```

---

```{r}
factorial(1500)
prod(seq(1, 1500))
```

```{r}
log(prod(seq(1, 1500)))
sum(log(1:1500))
```


---

```{r}
sum(log(1:1500)) - sum(log(1:1498)) - sum(log(1:2))
exp(sum(log(1:1500)) - sum(log(1:1498)) - sum(log(1:2)))
```

--

* Compare with

```{r}
log(choose(1500, 2))
choose(1500, 2)
```

---

layout: false

# R can draw graphs

```{r}
plot(x, p.x, ylab = "Probability", pch = 16)
title(main = sprintf("Binomial(%g, %g)", N, p))
abline(h = 0, col = "grey")
```

---

# R can simulate random variables

```{r}
cards <- as.vector(outer(c("H", "D", "C", "S"), 1:13, paste, sep = "-"))
cards
```

--

```{r}
sample(cards, 13)
sample(cards, 13)
```

---

layout: true

# Demonstration: The law of large numbers

---

* Does proportion of tosses that turn up head converge to probability?

--

```{r}
p <- 0.25
z <- rbinom(100, size = 1, prob = p)
z
```

--

```{r}
cumsum(z)
```

---

```{r}
cumsum(z) / 1:100
```

---

* How _fast_ does the sample proportion converge?

.scrollable500[

```{r,fig.height = 6}
plot(1:100, type = "n", ylim = c(0, 1), ylab = "Sample proportion")
for (i in 1:50) {
    z <- rbinom(100, size = 1, prob = p)
    lines(1:100, cumsum(z) / 1:100, col = sample(colors(), 1))
}
```

]





