---
layout: remark
title: Lazy Evaluation
subtitle: Data Analysis with R and Python
author: Deepayan Sarkar
mathjax: true
---

layout: true

# Lazy Evaluation

---

* Consider the following function

```{r opts, echo = FALSE, results = "hide", warning = FALSE, message = FALSE}
library(knitr)
library(kableExtra)
opts_chunk$set(cache = TRUE, cache.path='~/knitr-cache/darp-lazyeval/', autodep = TRUE,
               comment = "", warning = TRUE, message = TRUE,
               fig.width = 12, fig.height = 5.5,
               dev = "svglite", dev.args = list(pointsize = 15, bg = "transparent"),
               knitr.table.format = "html",
               fig.path='figures/4-lazyeval-')
options(warnPartialMatchDollar = FALSE, width = 90) # 110
```


<div>
$$
\newcommand{\sub}{_}
\newcommand{\nseq}[1]{ {#1}\sub{1}, {#1}\sub{2}, \dotsc, {#1}\sub{n} }
$$
</div>

```{r}
choose1 <- function(u, a, b) {
    if (u < 0.5) a else b
}
```

---

* Chooses and returns one of two arguments depending on a third argument

--

* Could be used for treatment randomization

```{r}
choose1(rbinom(1, size = 1, prob = 0.3), "placebo", "treatment")
choose1(rbinom(1, size = 1, prob = 0.3), "placebo", "treatment")
choose1(rbinom(1, size = 1, prob = 0.3), "placebo", "treatment")
choose1(rbinom(1, size = 1, prob = 0.3), "placebo", "treatment")
```

---

* How does this function work?

--

* Maybe

	* The function is called with some arguments
	
	* R matches and evaluates the arguments `u`, `a`, and `b`

	* Body of function executed with these values

--

* Does this sound reasonable?

* How can we check? [Code demo](lazy-evaluation-demo.R)

* What happens in Python?

---

layout: true

# Non-standard Evaluation

---

* Suppose we have the following problem: 

	* Find all subsets of size 2 from a set of size 5

---

* Easy to create all _ordered_ pairs using the `expand.grid()` function

.scrollable500[

```{r}
g <- expand.grid(a = 1:5, b = 1:5)
g
```

]

---

* But we want to retain only the $5 \choose 2$ distinct combinations of size 2

* Also not difficult: 

```{r}
g$b > g$a            # find the rows for which $a > b$
```

--

```{r}
g[g$b > g$a, ]       # Then use this as a row index
```

---

* As noted before, referring to `g` multiple times not ideal

* We can avoid doing this using the `eval()` function

---

* We can replace

```{r}
g$b > g$a
```

* by

```{r}
eval(quote(b > a), g)
```

--

* A general function that makes this even simpler is the `with()` function

```{r}
with(g, b > a)
```

---

* We might guess that this is done using `substitute()`

--

* We can verify this by looking at the `with` function by typing its name

```{r}
with
```

---

* Actually this just tells us that `with()` is a _generic_ function

* The _method_ we want is the default method

```{r}
getS3method("with", "default")
```

---

* All of this can now be combined to get all the $5 \choose 2$ distinct 2-subsets

```{r}
g[with(g, b > a), ]
```

---

* This is a very common use case: obtaining a subset a dataset

* There is an even simpler way using the `subset()` function

```{r}
subset(g, b > a)
```

---

* A popular add-on package __dplyr__ has a similar function called `filter()`

```{r}
dplyr::filter(g, b > a)
```

--

* Here `dplyr::filter` is the _namespace_ notation `package::function`

* Allows us to use the function without attaching the whole package

* Similar to the `.` accessor in Python


---

* The __dplyr__ package is very useful for routine data manipulation

* We will discuss it again later

* The idea of quoted expressions and _non-standard evaluation_ is
  essential in many other contexts






