---
layout: remark
title: Using ggplot2
subtitle: Data Analysis with R and Python
author: Deepayan Sarkar
mathjax: true
---

layout: true

# ggplot2

---

* Traditional graphics and lattice share a common philosophy

	* Plotting functions are named after plot type
	
	* Example: `graphics::hist()` and `lattice::histogram()`

--

* ggplot approaches data visualization in a different way


```{r opts, echo = FALSE, results = "hide", warning = FALSE, message = FALSE}
library(knitr)
library(kableExtra)
opts_chunk$set(cache = TRUE, cache.path='~/knitr-cache/darp-ggplot/', autodep = TRUE,
               comment = "", warning = TRUE, message = TRUE,
               fig.width = 12, fig.height = 5.5,
               dev = "svglite", dev.args = list(pointsize = 15, bg = "transparent"),
               knitr.table.format = "html",
               fig.path='figures/7-ggplot-')
options(warnPartialMatchDollar = FALSE, width = 90) # 110
```



<div>
$$
\newcommand{\sub}{_}
$$
</div>



---

* Want to recreate the following histogram using ggplot2

```{r}
data(NHANES, package = "NHANES")
lattice::histogram( ~ BPSysAve | Gender, data = NHANES,
                    layout = c(1, 2), nint = 25)
```

---

* The gg in ggplot stands for _grammar of graphics_

* Approach:

	* Define an abstract grammar for visualization

	* Consists of various predefined components

	* Plots are constructed by combining these components

---

* We start by specifying two of those components:

	* The dataset to be used

	* The faceting variable
	
```{r}
library(package = "ggplot2")
p <- ggplot(data = NHANES) + facet_wrap(~ Gender, ncol = 1)
```

---

* Note that we have assigned the result to a variable

--

* This is an important feature of __lattice__ and __ggplot__

	* Traditional graphics calls plot something as soon as they are called

	* This is not true for __lattice__ and __ggplot__
	
	* Instead, they create objects that contain the information required for a plot

---

* Actual plot is created when they are 'printed' (blank because it doesn't have any geoms)

```{r}
p
```

---

* Fortunately, there is a "histogram" geom  we can use

```{r}
p + geom_histogram(aes(x = BPSysAve), bins = 25)
```

---

layout: true

# Stats

---

* This approach hides an intermediate step

* We map average systolic blood pressure to the x-coordinate (aesthetic mapping)

* However, the y-coordinate mapping is not explicitly mapped

--

* What is actually plotted here are summary statistics (bin counts)

* Even the x-coordinate is no longer the raw data, but only the bin locations

--

* So essentially raw data has been summarized to create some new data (which is plotted)

* Such statistical summaries are called __"stats"__ in ggplot2 terminology

---

* We can make this operation explicit as follows

```{r}
p + stat_bin(aes(x = BPSysAve), bins = 25, na.rm = TRUE)
```

---

* This 'stat' is applied by default in `geom_histogram()`

```{r}
str(geom_histogram)
```

--

* Similarly, `stat_bin()` has a default 'geom'

```{r}
str(stat_bin)
```

---

* Thus, the last plot we created is actually

```{r}
p + stat_bin(aes(x = BPSysAve), bins = 25, geom = "bar", na.rm = TRUE)
```

---

* We can change the geom to get a different visualization of the same summary

```{r}
p + stat_bin(aes(x = BPSysAve), bins = 25, geom = "point", na.rm = TRUE)
```

---

* A more useful choice of geom is "line" (result is a reasonable substitute for histograms)

```{r}
p + stat_bin(aes(x = BPSysAve), bins = 25, geom = "line", na.rm = TRUE)
```

---

* These plots show bin counts

* As we saw earlier, histograms can also be viewed as _density_ estimators

* Read help page of `stat_bin()`: it also computes estimated density

---

* For these, y-axis mapping must be `density`, not `count`

```{r}
p + stat_bin(aes(x = BPSysAve, y = after_stat(density)), bins = 25, geom = "line", na.rm = TRUE)
```

---

* Alternative method: average shifted histogram / kernel density estimation 

```{r}
p + stat_density(aes(x = BPSysAve), na.rm = TRUE)
```

---

* Again, replacing the default geom ("area") by "line" gives more familiar density plot

```{r}
p + stat_density(aes(x = BPSysAve), geom = "line", na.rm = TRUE)
```

---

layout: true

# Aesthetic mapping

---

* Important feature of ggplot2: _aesthetic mapping_

	* Aesthetic attributes of any geom can be mapped to specific variables
	
	* This includes, color, fill, size, shape, etc.
	
	* Details depend on specific geom

--

```{r}
str(geom_line)
```

---

* Example: map `Race1` (categorical variable) to line color

```{r}
p + stat_density(aes(x = BPSysAve, color = Race1), geom = "line", na.rm = TRUE)
```

---

* Example: Default `position = "stack"` is useful for default geom "area"

```{r}
p + stat_density(aes(x = BPSysAve, fill = Race1), geom = "area", na.rm = TRUE)
```

---

```{r}
str(stat_density)
```

---

```{r}
p + stat_density(aes(x = BPSysAve, color = Race1),
                 geom = "line", position = "identity", na.rm = TRUE)
```


---

* The idea of aesthetic mapping of variables to visual attributes is very powerful

* We will demonstrate this with the gapminder data

```{r}
library(gapminder)
str(gapminder)
```

---

* Create a blank ggplot object with data and a faceting variable (year)

```{r}
pgm <- ggplot(data = gapminder) + facet_wrap(~ year)
```

---

* Scatter plot to study relation between `lifeExp` and `gdpPercap`

```{r}
pgm + geom_point(aes(x = gdpPercap, y = lifeExp))
```


---

* `gdpPercap` is highly skewed, so take logarithm

```{r}
pgm + geom_point(aes(x = log(gdpPercap), y = lifeExp))
```

---

* Other attributes: map continent to color

```{r}
pgm + geom_point(aes(x = log(gdpPercap), y = lifeExp, color = continent))
```


---

* Another continuous variable is population: map to _size_ of points (bubble chart)

```{r}
pgm + geom_point(aes(x = log(gdpPercap), y = lifeExp, color = continent, size = pop))
```


<!-- # Summary -->

<!-- # Interactive graphics from static ggplot plots -->



