---
layout: remark
title: Data Visualization — Summary and Further Topics
subtitle: Data Analysis with R and Python
author: Deepayan Sarkar
mathjax: true
---

layout: true

# Traditional graphics

---

* Implemented in the  __graphics__ package

* Add-on packages provide further functionality

* Mainly two types

	- High-level functions

	- Low-level functions


```{r opts, echo = FALSE, results = "hide", warning = FALSE, message = FALSE}
library(knitr)
opts_chunk$set(cache = TRUE, cache.path='~/knitr-cache/darp-viz-summary/', autodep = TRUE,
               comment = "", warning = TRUE, message = TRUE,
               fig.width = 12, fig.height = 5.5,
               dev = "svglite", dev.args = list(pointsize = 15, bg = "transparent"),
               knitr.table.format = "html",
               fig.path='figures/8-vizsummary-')
options(warnPartialMatchDollar = FALSE, width = 90) # 110
```



<div>
$$
\newcommand{\sub}{_}
$$
</div>

---

layout: false

# Common High-Level functions in __graphics__


| Function         | Default Display                                   |
|:-----------------|:--------------------------------------------------|
| `stripchart()`   | Strip Chart (Comparative 1-D Scatter Plots)       |
| `boxplot()`      | Comparative Box-and-Whisker Plots                 |
| `hist()`         | Histogram                                         |
| `plot(density())`| Kernel Density Plot                               |
| `plot()`         | Scatter Plot, Time-series Plot (with `type="l"`)  |
| `barplot()`      | Bar Plot                                          |
| `dotchart()`     | Cleveland Dot Plot                                |
| `qqnorm()`       | Normal Quantile-Quantile Plot                     |
| `qqplot()`       | Two-sample Quantile-Quantile Plot                 |
| `pairs()`        | Scatter-Plot Matrix                               |


---

# Common High-Level functions in __lattice__

| Function         | Default Display                                   |
|:-----------------|:--------------------------------------------------|
| `stripplot()`    | Strip Chart (Comparative 1-D Scatter Plots)       |
| `bwplot()`       | Comparative Box-and-Whisker Plots                 |
| `histogram()`    | Histogram                                         |
| `densityplot()`  | Kernel Density Plot                               |
| `xyplot()`       | Scatter Plot, Time-series Plot (with `type="l"`)  |
| `barchart()`     | Bar Plot                                          |
| `dotplot()`      | Cleveland Dot Plot                                |
| `qqmath()`       | Normal Quantile-Quantile Plot                     |
| `qq()`           | Two-sample Quantile-Quantile Plot                 |
| `splom()`        | Scatter-Plot Matrix                               |


---

# Some useful low-level functions in __graphics__

| Function      | Purpose                               |
|:--------------|:--------------------------------------|
| `text()`      | Add Text to a Plot                    |
| `lines()`     | Add Connected Line Segments to a Plot |
| `points()`    | Add Points to a Plot                  |
| `polygon()`   | Add Polygons to a Plot                |
| `rect()`      | Add Rectangles to a Plot              |
| `segments()`  | Add Line Segments to a Plot           |


---

# Some other low-level utility functions in __graphics__

| Function     | Purpose                            |
|:-------------|:-----------------------------------|
| `abline()`   | Add Straight Lines to a Plot       |
| `arrows()`   | Add Arrows to a Plot               |
| `axis()`     | Add an Axis to a Plot              |
| `box()`      | Draw a Box around a Plot           |
| `grid()`     | Add Grid to a Plot                 |
| `legend()`   | Add Legends to Plots               |
| `title()`    | Add Annotation (Titles and Labels) |


---

# Common low-level functions in __lattice__

| Function           | Purpose                                |
|:-------------------|:---------------------------------------|
| `panel.text()`     | Add Text to a Panel                    |
| `panel.lines()`    | Add Connected Line Segments to a Panel |
| `panel.points()`   | Add Points to a Panel                  |
| `panel.polygon()`  | Add Polygons to a Panel                |
| `panel.rect()`     | Add Rectangles to a Panel              |
| `panel.segments()` | Add Line Segments to a Panel           |
| `panel.abline()`   | Add Straight Lines to a Panel          |
| `panel.arrows()`   | Add Arrows to a Panel                  |
| `panel.grid()`     | Add Grid to a Panel                    |


---

# The ggplot2 approach

* Specify a plot using a __grammar__ consisting of

	* Geometric constructs

	* Aesthetic Mappings

	* Statistical Summaries

---

# ggplot2: List of geoms

```{r}
library(ggplot2)
apropos("^geom_")
```


???

---

# ggplot2: List of stats

```{r}
apropos("^stat_")
```

---

# Interactive Graphics

* R works on many platforms: Windows, Mac, Linux, UNIX

--

* Static graphics achieves portability through _graphics devices_

--

* Cross-platform interactive graphics is even more difficult

---

# Interactive Graphics: Cross-platform technology

- OpenGL (R package __rgl__)

- Browser-based visualization technology (R packages __plotly__, __rbokeh__)

---

layout: true

# Plotly: Example

---

```{r,message=FALSE}
library(package = "ggplot2")
library(package = "plotly")
library(package = "gapminder")
p1 <- 
    ggplot(data = gapminder,
           mapping = aes(x = log(gdpPercap), y = lifeExp,
                         size = pop, color = continent)) +
    facet_wrap(~ year) + geom_point()
```

---

```{r}
p1
```

---

```r
ggplotly(p1)
```


```{r,echo=FALSE}
library(htmlwidgets)
ggplotly(p1) |> saveWidget(file = "figures/8-gapminder-plotly-1.html",
                           libdir = "plotly", selfcontained = FALSE)
```

<iframe src="figures/8-gapminder-plotly-1.html" 
        style="border: none;"
        width="100%" height="500px" />


---

* Want to add country name

* Easiest way is to add a fake aesthetic


```{r}
p2 <- 
    ggplot(data = gapminder,
           mapping = aes(x = log(gdpPercap), y = lifeExp,
                         size = pop, color = continent,
                         tooltip = country)) +
    facet_wrap(~ year) + geom_point()
```

---

```r
ggplotly(p2)
```


```{r,echo=FALSE}
library(htmlwidgets)
ggplotly(p2) |> saveWidget(file = "figures/8-gapminder-plotly-2.html",
                           libdir = "plotly", selfcontained = FALSE)
```

<iframe src="figures/8-gapminder-plotly-2.html" 
        style="border: none;"
        width="100%" height="500px" />


