---
layout: remark
title: Distributions
subtitle: Statistics I — Data Exploration
author: Deepayan Sarkar
mathjax: true
---


```{r opts, echo = FALSE, results = "hide", warning = FALSE, message = FALSE}
opts_chunk$set(cache = TRUE, cache.path='~/knitr-cache/de-distribution/', autodep = TRUE,
               comment = "", warning = TRUE, message = TRUE,
			   ## engine.path = list(python = "/usr/bin/python3"),
               fig.width = 12, fig.height = 5,
               dev = "svglite", dev.args = list(pointsize = 12),
               knitr.table.format = "html",
			   fig.path='figures/distribution-')
options(warnPartialMatchDollar = FALSE, width = 80)
suppressWarnings(require(dplyr))
```



<div>
$$
\newcommand{\sub}{_}
$$
</div>


# Goals

* Some preliminary concepts

	* Visualization

	* Probability distributions

	* Data Frame (R equivalent of a spreadsheet)

---

layout: true

# Probability distribution 

---

* Example: _Number of distinct birthdays_ in a group of $n$ people

* This is different from the earlier problem (probability of all distinct)

--

* One way to simulate:

```{r}
distinct_birthdays <- function(n) {
    sample(1:365, n, replace = TRUE) |> unique() |> length()
}
```

* The `|>` notation represents a "pipe" operator

* This is equivalent to 

```r
distinct_birthdays <- function(n) {
    length(unique(sample(1:365, n, replace = TRUE)))
}
```

---

* This is a 'random' quantity — depends on the sample drawn

--

* We can repeat the simulation experiment and obtain 'data'

```{r}
d <- replicate(25, distinct_birthdays(23))
d # each element represents result of one simulation
```

--

* How can we plot this data? 

* Do we even want to?

---

layout: true

# The `plot()` function

---

```{r}
plot(d) # not very useful in this case
```

---

```{r}
plot(sort(d), pch = 16) # sorting makes it slightly more useful 
```

---

layout: true

# The `stripchart()` function

---

```{r}
stripchart(d) # also mostly useless
```

---

```{r}
stripchart(d, method = "jitter") # but stripchart() has some useful options
```

---

```{r}
stripchart(d, method = "stack") # more useful: the "stack" method
```

---

* Let's try with a larger $n$ and more replications

```{r}
d50 <- replicate(100, distinct_birthdays(50))
d50
```

---

```{r}
stripchart(d50, method = "stack", pch = 16)
```

---

layout: true

# Tables: `barplot()` and `dotchart()`

---

* By stacking, we are basically counting the "frequency" of each value

--

* This can be done directly using the `table()` function

```{r}
table(d50)
```

---

* We are actually interested in just the proportion

```{r}
table(d50) |> prop.table()
```

* Tables can be plotted using "bar charts" and "dot plots"

---

```{r}
table(d50) |> prop.table() |> barplot()
```

---

```{r, warning = FALSE}
table(d50) |> prop.table() |> dotchart(pch = 16)
```

---

layout: true

# Alternative functions: `barchart()` and `dotplot()`

---

* Did you notice that the names are switched around?

--

* R comes with an add-on package called `lattice` which has alternative implementations

* To use it, you need to "load" the `lattice` package

```{r}
library(package = "lattice")
```

---

```{r}
table(d50) |> prop.table() |>
    barchart()
```

---

```{r}
table(d50) |> prop.table() |>
    barchart(horizontal = FALSE)
```

---

```{r}
replicate(100, distinct_birthdays(50)) |> table() |> prop.table() |>
    barchart(horizontal = FALSE) # another simulation
```

---

```{r}
replicate(100000, distinct_birthdays(50)) |> table() |> prop.table() |>
    barchart(horizontal = FALSE) # increase replications
```

---

```{r}
replicate(100000, distinct_birthdays(60)) |> table() |> prop.table() |>
    barchart(horizontal = FALSE) # n = 60
```

---

```{r}
replicate(100000, distinct_birthdays(70)) |> table() |> prop.table() |>
    barchart(horizontal = FALSE) # n = 70
```

---

```{r}
replicate(100000, distinct_birthdays(80)) |> table() |> prop.table() |>
    barchart(horizontal = FALSE) # n = 80
```

---

```{r}
replicate(100000, distinct_birthdays(80)) |> table() |> prop.table() |>
    dotplot(horizontal = FALSE) # n = 70
```

---

layout: true

# Distribution

---

* What we are really trying to visualize is a "probability distribution"

--

* Let $X$ be the number of distinct birthdays in a group of 80 people

	* What is $P(X = 80)$ ?

--

	* What is $P(X = 79)$ ?

	* What is $P(X = 78)$ ?

--

	* What is the most likely value of $X$ ?

	* What is the expected value of $X$ ?

--

* In this case, true answer is difficult to calculate

* Simulation will give us answers that will be close (by "law of large numbers")


---

layout: true

# The Binomial distribution

---

* An important distribution: number of successes in $n$ independent trials (e.g., coin tosses)

* The probabily of success in one trial is $p$ (which is known, but not necessarily $\frac12$)

--

* We can simulate exactly as before

```{r}
n <- 80
sample(c(0, 1), size = n, replace = TRUE)
sample(c(0, 1), size = n, replace = TRUE) |> sum()
```

---

```{r}
replicate(100000, sum(sample(c(0, 1), size = 80, replace = TRUE))) |>
    table()
```

---

```{r}
replicate(100000, sum(sample(c(0, 1), size = 80, replace = TRUE))) |>
    table() |> prop.table()
```

---

```{r}
replicate(100000, sum(sample(c(0, 1), size = 80, replace = TRUE))) |> 
    table() |> prop.table() |> barchart(horizontal = FALSE)
```

---

```{r}
replicate(100000, sum(sample(c(0, 1), size = 80, replace = TRUE, prob = c(0.1, 0.9)))) |> 
    table() |> prop.table() |> barchart(horizontal = FALSE)
```

---

* In this example, the exact probabilities are actually easy to calculate

* Let $X$ be the number of successes in $n$ independent trials with probabily of success $p$. Then, 

$$
P(X = k) = { n \choose k } \, p^k \, (1-p)^{n - k}\,, \, k = 0, 1, 2, \dotsc, n
$$

---

* To compare, we need both simulation probabilities and exact probabilities

* This is where data frames become useful

--

* Data frames are like a spreadsheet, containing multiple columns

* Every column must have a name, and must have the same length

---

* Example: Proportions observed in simulation

```{r}
X <- replicate(100000, sum(sample(c(0, 1), size = 80, replace = TRUE, prob = c(0.1, 0.9))))
T <- prop.table(table(X))
```

--

```{r}
T
str(T)
```


---

.scrollable500[

```{r}
df <- data.frame(k = as.integer(names(T)), proportion = as.numeric(T))
df
```

]

---

.scrollable500[

```{r}
df$probability <- with(df, choose(80, k) * 0.9^k * 0.1^(80-k))
df
```

]

---

```{r}
with(df, plot(k, proportion, pch = 16))
with(df, lines(k, probability, col = "blue"))
```

---

layout: false
class: center, middle

# Questions?





