---
layout: remark
title: Data Collection and Summarization
subtitle: Statistics I — Data Exploration
author: Deepayan Sarkar
mathjax: true
---


```{r opts, echo = FALSE, results = "hide", warning = FALSE, message = FALSE}
opts_chunk$set(cache = TRUE, cache.path='~/knitr-cache/de-data-summary/', autodep = TRUE,
               comment = "", warning = TRUE, message = TRUE,
			   ## engine.path = list(python = "/usr/bin/python3"),
               fig.width = 12, fig.height = 5,
               dev = "svglite", dev.args = list(pointsize = 12),
               knitr.table.format = "html",
			   fig.path='figures/data-summary-')
options(warnPartialMatchDollar = FALSE, width = 80)
suppressWarnings(require(lattice))
```



<div>
$$
\newcommand{\sub}{_}
\newcommand{\set}[1]{\left\lbrace {#1} \right\rbrace}
\newcommand{\nseq}[1]{ {#1}\sub{1}, {#1}\sub{2}, \dotsc, {#1}\sub{n} }
$$
</div>


# Goals

* Where do data come from?

* Can we classify datasets by how they were collected?

* How can we summarize data numerically or graphically?

---

layout: true

# Sources of data

---

* Traditional data types

	* Categorical - nominal / ordered

	* Numeric - discrete / continuous

--

* Data "modes" that more difficult to analyse (but are of increasing interest)

	* Free text
	
	* Images
	
	* Sound
	
	* Many others
	
--

* But how are such data typically collected?

---

* Key methodologies

	* Census
	
	* Sample survey
	
	* Observational studies / Case-control studies
	
	* Randomized studies / Randomized controlled trials


* Key concepts

	* Observational units

	* Population
	
	* Sample
	
---

layout: true

# Some scenarios

---

* We will consider some specific scenarios along with a question of interest

* In each case, we want to plan a data collection experiment that will answer the question 

---

* What is the leading cause of death for adults in India?

	* Does it vary by sex?
	
	* Does it vary by state? By district?
	
	* Does it vary by education level? By income?
	
	* Do the answers change over time?
	
--

* What is the target population?

* Can we do a census?

* Can we do a survey?

* How should we sample?

* What are the observational units?

* Observational or randomized? Controls?

---

* How much money does a typical adult in India make?

	* Does it vary by sex?
	
	* Does it vary by state? By district?
	
	* Does it vary by education level? By employment status?
	
	* Do the answers change over time?
	
--

* What is the target population?

* Can we do a census?

* Can we do a survey?

* How should we sample?

* What are the observational units?

* Observational or randomized? Controls?

---

* Consider a "random" person who will be born in India in 2025

* How likely is it that they will be born on January 1? January 2? ... December 31?

--

* What is the target population?

* Can we do a census?

* Can we do a survey?

* How should we sample?

* What are the observational units?

* Observational or randomized? Controls?

--

* Here is a [relevant dataset](https://www.ons.gov.uk/visualisations/nesscontent/dvc307/line_chart/data.csv)  (CSV) from [England and Wales](https://www.ons.gov.uk/peoplepopulationandcommunity/birthsdeathsandmarriages/livebirths/articles/howpopularisyourbirthday/2015-12-18)

* Can we use this data to answer the same question for England and Wales?

--

* Similar but more important question: Predict the __air quality__ in Delhi from October to December?

---

* A medicine factory produces tablets that are supposed to have a certain chemical composition

	* A regulator wants to check whether the actual composition is within acceptable limits

	* However, the only way to test a tablet is "destructive", i.e., the tablet cannot be used afterward

--

* What is the target population?

* Can we do a census?

* Can we do a survey?

* How should we sample?

* What are the observational units?

* Observational or randomized? Controls?

---

* A new yearly injection can potentially delay the onset of diabetes

	* However, both the amount of benefit and the possible side effects are currently unknown

	* We want to plan an experiment to study both

--

* What is the target population?

* Can we do a census?

* Can we do a survey?

* How should we sample?

* What are the observational units?

* Observational or randomized? Controls?

---

* Your friend loves a particular type of pizza from Dominos

* He also claims he can differentiate between the pizzas from two different outlets in your neighbourhood

* You want to test whether this claim is true

--

* What is the target population?

* Can we do a census?

* Can we do a survey?

* How should we sample?

* What are the observational units?

* Observational or randomized? Controls?

---

* What is the GDP of India? What is the average life expectancy in India?

* How do they compare with other contries?

* How have they changed over time?

--

* How do we measure?

* More basic question: how are they defined?

--

* These are by definition __summary__ measures

* Example: [TSV data](https://deepayan.github.io/BSDS/2024-01-DE/data/gapminder.tsv) and [visualization](https://www.gapminder.org/tools/) from GapMinder

* The actual process to calculate these from unit-level data is probably quite complicated


---

layout: true

# What do we actually do with data?

---

* The process of data analysis is often involves just calculating various summary measures

--

* In fact, the technical name for a summary measure computed from data is a __statistic__

* In this course, we will mostly learn about commonly used summary statistics

--

* But also important to remember that summary statistics are usually computed from _samples_

* We need to be careful when we use them to make conclusions about a larger _population_

---

* Example: Are birthdays equally likely? 

* Let $X$ be the smallest birth month frequency in a 'random sample' of $n = 65$ people

* We take one sample, where we observe $X = 3$. What can we conclude?

--

```{r}
min_count <- function(n, pmonths = NULL)
{
    b <- sample(1:12, n, prob = pmonths, replace = TRUE)
    T <- table(b)
    if (length(T) < 12) {
        return(0)
    }
    else {
        return(min(T))
    }
}
```

---

```{r}
replicate(10000, min_count(65)) |>
    table() |> prop.table() |> barplot()
```


---

```{r}
replicate(10000, min_count(65, pmonths = c(31, 28, 31, 30, 31, 30, 31, 31, 30, 31, 30, 31) / 365)) |>
    table() |> prop.table() |> barplot()
```

---

* Example: Are birthdays equally likely? 

* Let $X$ be the smallest birth month frequency in a 'random sample' of $n = 65$ people

* We take one sample, where we observe $X = 3$. What can we conclude?

--

* We __cannot__ conclude that birthdays are _not_ equally likely

---

layout: true

# What statistics are useful?

---

* This is a natural and important question

* The answer depends on the type of data and the problem of interest

---

* How can we summarize categorical variables such as _cause of death_?

--

	* We usually want to find the probabilities of various categories

	* Natural summary statistic is sample proportion
	
	* The 'most likely' category is known as the __mode__

---

* How can we summarize numeric variables such as _income_?

	* This is much more difficult to answer
	
	* Common summary statistics: __mean__ and __median__
	
	* But these capture only limited aspects of the distribution
	
--

* We will learn about mean, median and similar summary statistics

* But first we will learn about the _empirical distribution_


---

layout: true

# Empirical distribution

---

* Suppose we have observed values $\nseq{X}$ of some variable

* Assume that the values of $\nseq{X}$ are all _distinct_

* The empirical distribution is a probability distribution whose

	* Sample space is the set $\set{ \nseq{X} }$
	
	* All elements in the sample space are equally likely (has probability $1/n$)
	
--

* This is adjusted suitably if values are not distinct

	* The sample space consists of all distinct values

	* If the $i$th value appears $k\sub{i}$ times, the corresponding probability if $k\sub{i} / n$

---

* The empirical distribution contains all information in the data...

	* as long as the observations are "independent" (a 'random sample')
	
* As it is a probability distribution, we can apply the tools we have to study distributions

--

* Why is the empirical distribution important?

	* We have seen that sample proportions 'converge' to probabilities as sample size $n$ increases

	* More generally, the empirical distribution also 'converges' to the population distribution

--
	
	* This is one of the _main justifications_ for using observed data to 'infer' about underlying population

--

	* This convergence is mathematically more complicated (which will not be discussed in this course)
	
	* However, understanding this convergence graphically is important 

---

layout: true

# Visualizing the empirical distribution

---

* Visualizing the empirical distribution is important but challenging

--

* Let us consider the height data we collected in our survey

```{r}
survey <- read.csv("https://deepayan.github.io/BSDS/2024-01-DE/data/bsds-survey.csv")
height <- survey$height
height
```

* How do we visualize it?

---

```{r}
stripchart(height, method = "stack")
```

---

* For comparison, let us simulate data uniformly between 150 and 200

```{r}
n <- length(height)
u <- sample(150:195, n, replace = TRUE)
stripchart(list(unif = u, obs = height), method = "stack")
```

---

* Are the two samples similar or very different? Not very easy to say

--

* Another possible way to compare

```{r}
par(mfrow = c(1, 2)); plot(u,       ylim = c(150, 195)); plot(height,       ylim = c(150, 195))
```

---

* There seems to be some qualitative difference

* This difference becomes clearer if we plot _sorted_ data

```{r}
par(mfrow = c(1, 2)); plot(sort(u), ylim = c(150, 195)); plot(sort(height), ylim = c(150, 195))
```

---

* It is more traditional to swap the axes and convert the y-axis to a probability scale

```{r}
par(mfrow = c(1, 2)); plot(sort(u), ppoints(n), xlim = c(150, 195));
                      plot(sort(height), ppoints(n), xlim = c(150, 195))
```

---

* This is equivalent to the _empirical cumulative distribution function_ (ECDF)

```{r}
par(mfrow = c(1, 2)); plot(ecdf(u), xlim = c(150, 195));
                      plot(ecdf(height), xlim = c(150, 195))
```

---

layout: true

# The empirical cumulative distribution function (ECDF)

---

* Height data from a different source (the NHANES survey)

```{r}
library(NHANES)
dim(NHANES)
nhsub1 <- subset(NHANES, !is.na(Height)) # remove missing height values
n1 <- nrow(nhsub1)
n1
```

---

* ECDF of height of a random sample of Americans

```{r}
plot(sort(nhsub1$Height), ppoints(n1), type = "l")
```

---

* Low heights are from children

* We are interested in the height distribution among adults

```{r}
nhsub2 <- subset(nhsub1, Age >= 20) # remove children
n2 <- nrow(nhsub2)
n2
```

---

* ECDF of height of a random sample of American adults

```{r}
plot(sort(nhsub2$Height), ppoints(n2), type = "l")
```

---

* ECDF of the magic distribution: Normal

```{r}
plot(sort(rnorm(n2)), ppoints(n2), type = "l")
```

---

layout: true

# Quantile-Quantile plots

---

* These ECDF plots essentially show sorted data vs equally spaced probabilities

--

* Useful alternative: plot sorted data against sorted data from the 'Normal distribution'

```{r}
par(mfrow = c(1, 2)); plot(sort(rnorm(n2)), sort(nhsub2$Height))
```

---

* How do we get data from the 'Normal distribution'? Simulate using `rnorm()`

* Problem: Every time we simulate, we will get a slightly different plot

```{r}
par(mfrow = c(1, 2)); plot(sort(rnorm(n2)), sort(nhsub2$Height));
```

---

* Solution: Instead of simulating, use something called 'theoretical quantiles' of the Normal distribution

* The result is known as a _Normal Quantile-Quantile Plot_ or simply _Normal Q-Q plot_

```{r}
par(mfrow = c(1, 2)); plot(sort(rnorm(n2)), sort(nhsub2$Height)); qqnorm(nhsub2$Height)
```

---

* As before, the lattice package has a different implementation that is more powerful

```{r}
qqmath(~ Height, data = nhsub2, grid = TRUE, aspect = 1, groups = Gender, auto.key = TRUE)
```

---

layout: true

# Questions

---

* This raises several important questions:

	* What is the Normal distribution and why should we compare with it?

	* What are quantiles?

	* What are theoretical quantiles?

	* What can Q-Q plots tell us?
	
--

* The first two questions will be discussed in your probability course

* We will give very vague answers for now

---

* What is the Normal distribution

	* The Normal distribution is a very simple _continuous_ distribution
	
	* It is best for now to think of it as an _approximation_ to various distributions that we observe in real life

	* We will see examples of this soon

---

* What are quantiles?

	* In Q-Q plots we plot sorted data
	
	* Sorted data have a special name in statistics: _order statistics_
	
--

	* Specifically, the $k$-th value of $X\sub{i}$ in sorted order is known as the $k$-th order statistic
	
	* The standard notation for the $k$-th order statistic is $X\sub{(k)}$

--

	* Roughly speaking, quantiles are order statistics, but identified by their _relative rank_ (like percentiles)
	
---

* Examples of quantiles:

	* The $90$-th percentile is the same as the $0.9$ quantile
	
	* This is defined as the number $Q$ such that 90% of the data is less than (or equal to) $Q$
	
	* In a sample of size $n = 100$, this would be $X\sub{(90)}$

	* In a sample of size $n = 1000$, this would be $X\sub{(900)}$

--

	* In a sample of size $n = 7182$, this would be ?? 

--

```{r}
0.9 * 7182
```

---
	
* Several approximations available; see `help(quantile)`
	
```{r}
with(subset(nhsub2, Gender == "male"), quantile(Height, probs = c(0.2, 0.4, 0.6, 0.8)))
with(subset(nhsub2, Gender == "female"), quantile(Height, probs = c(0.2, 0.4, 0.6, 0.8)))
```

---

* What are theoretical quantiles?

	* Similar idea, but with probability instead of relative rank
	
	* $p$-th quantile is the number $Q$ such that $P(X \leq Q) = p$
	
	* Needs to be modified suitably to account for discrete jumps
	
--

* R has _quantile functions_ for all standard distributions

```{r}
qbinom(c(0.2, 0.4, 0.6, 0.8), size = 20, prob = 0.25)
qbinom(c(0.2, 0.4, 0.6, 0.8), size = 200, prob = 0.25)
qnorm(c(0.2, 0.4, 0.6, 0.8))
```

---

* Why are Q-Q plots useful?

	* Sample quantiles converge to theoretical quantiles of underlying population
	
	* Q-Q plots compare order statistics with corresponding theoretical quantiles 

--

* Why not ECDF?

	* This has to do with _human perception_
	
	* We find it easier to detect departures from a straight line (as opposed to a curve)

--
	
	* In principle, the theoretical quantiles can be from any appropriate distribution

	* The Normal distribution has been found to be the most useful default choice

---

* Two final points before moving on from Q-Q plots

	* In what sense do sample quantiles converge to population quantiles?
	
	* Why is the Normal distribution a useful reference for Q-Q plots?

* We will illustrate both these through simulation

* Theoretical answers will be discussed later in other courses

---

layout: true

# Convergence of sample quantiles

---

* Suppose data are generated from the Normal distribution

* Ideally, Q-Q plot should look like a straight line

* However, this will rarely happen for small sample sizes

---

```{r}
library(latticeExtra)
c(qqmath(~ rnorm(  10)), qqmath(~ rnorm(  10)), qqmath(~ rnorm(  10))) |>
    update(grid = TRUE, aspect = 1)
```

---

```{r}
c(qqmath(~ rnorm(  50)), qqmath(~ rnorm(  50)), qqmath(~ rnorm(  50))) |>
    update(grid = TRUE, aspect = 1)
```

---

```{r}
c(qqmath(~ rnorm( 100)), qqmath(~ rnorm( 100)), qqmath(~ rnorm( 100))) |>
    update(grid = TRUE, aspect = 1)
```


---

```{r}
c(qqmath(~ rnorm( 500)), qqmath(~ rnorm( 500)), qqmath(~ rnorm( 500))) |>
    update(grid = TRUE, aspect = 1)
```


---

```{r}
c(qqmath(~ rnorm(5000)), qqmath(~ rnorm(5000)), qqmath(~ rnorm(5000))) |>
    update(grid = TRUE, aspect = 1)
```

---

* Exercise: 

	* Try this for more sample sizes

	* Repeat several times to understand variability (as function of sample size)

--

* Q-Q plots can compare with quantiles of other distributions as well

* Let's try with Binomial

---

```{r}
size <- 100; prob <- 0.25; x <- rbinom(  10, size = size, prob = prob)
qqmath(~ x, distribution = function(p) qbinom(p, size = size, prob = prob),
       grid = TRUE, aspect = 1)
```

---

```{r}
size <- 100; prob <- 0.25; x <- rbinom(  50, size = size, prob = prob)
qqmath(~ x, distribution = function(p) qbinom(p, size = size, prob = prob),
       grid = TRUE, aspect = 1)
```

---

```{r}
size <- 100; prob <- 0.25; x <- rbinom( 100, size = size, prob = prob)
qqmath(~ x, distribution = function(p) qbinom(p, size = size, prob = prob),
       grid = TRUE, aspect = 1)
```

---

```{r}
size <- 100; prob <- 0.25; x <- rbinom( 500, size = size, prob = prob)
qqmath(~ x, distribution = function(p) qbinom(p, size = size, prob = prob),
       grid = TRUE, aspect = 1)
```

---

```{r}
size <- 100; prob <- 0.25; x <- rbinom(5000, size = size, prob = prob)
qqmath(~ x, distribution = function(p) qbinom(p, size = size, prob = prob),
       grid = TRUE, aspect = 1)
```


---

* So similar pattern: 

	* more variable for low sample size
	
	* Converges to straight line as sample size increases

--

* But what happens if we compare with Normal quantiles instead of Binomial quantiles?

---

```{r}
size <- 100; prob <- 0.25; x <- rbinom(  10, size = size, prob = prob)
qqmath(~ x, grid = TRUE, aspect = 1)
```

---

```{r}
size <- 100; prob <- 0.25; x <- rbinom(  50, size = size, prob = prob)
qqmath(~ x, grid = TRUE, aspect = 1)
```

---

```{r}
size <- 100; prob <- 0.25; x <- rbinom( 100, size = size, prob = prob)
qqmath(~ x, grid = TRUE, aspect = 1)
```

---

```{r}
size <- 100; prob <- 0.25; x <- rbinom( 500, size = size, prob = prob)
qqmath(~ x, grid = TRUE, aspect = 1)
```

---

```{r}
size <- 100; prob <- 0.25; x <- rbinom(5000, size = size, prob = prob)
qqmath(~ x, grid = TRUE, aspect = 1)
```

---

* This suggests that the Binomial$(n = 100, p = 0.25)$ distribution is well approximated by Normal

--

* This approximation improves with the number of Bernoulli trials $n$

* Let's try Binomial$(n = 1000, p = 0.25)$ 

---

```{r}
size <- 1000; prob <- 0.25; x <- rbinom(5000, size = size, prob = prob)
qqmath(~ x, grid = TRUE, aspect = 1)
```

---

* But the approximation is not as good for, say, Binomial$(n = 1000, p = 0.0025)$

```{r}
size <- 1000; prob <- 0.0025; x <- rbinom(5000, size = size, prob = prob)
qqmath(~ x, grid = TRUE, aspect = 1)
```

---

* Comparing to the true distribution will lead to a better visual fit

```{r}
qqmath(~ x, grid = TRUE, aspect = 1,
       distribution = function(p) qbinom(p, size = size, prob = prob))
```

---

* Summary

	* Sample quantiles (order statistics) converge to "theoretical quantiles" as sample size increases
	
	* Theoretical quantiles essentially determine underlying population

--

	* This means that looking at sample quantiles (through ECDF) tells us about underlying population
	
	* Q-Q plots transform ECDF to compare with a reference distribution (usually Normal)
	
--

	* The Normal distribution is a good reference in a surprisingly wide range of situations
	
	* However, there are also many situations where it is not appropriate
	
	* Systematic patterns in the Normal Q-Q plot can suggest alternative distributions (later)


---

layout: true

# Comparing multiple samples

---

* Q-Q plots are typically used to compare a sample to a reference distribution

* We are usually more interested in comparing two or more samples

---

* This is possible, as we saw before: compare heights of males and females in NHANES data

```{r}
qqmath(~ Height, data = NHANES, grid = TRUE, aspect = 1, groups = Gender, auto.key = TRUE, subset = Age >= 20)
```

---

* The difference can be summarized as a _constant_ shift in all the quantiles

* This can also be seen in a smaller subset of quantiles:

```{r}
s <- with(subset(NHANES, Age >= 20), split(Height, Gender))
str(s)
lapply(s, quantile, probs = c(0.2, 0.4, 0.6, 0.8),  na.rm = TRUE)
```

---

* It is traditional to summarize a sample with a specific set of __five__ quantiles

* This in known as the [__five number summary__](https://en.wikipedia.org/wiki/Five-number_summary), and can be produced by `fivenum()`

```{r}
lapply(s, fivenum)
```

---

* These are actually just the quantiles corresponding to probabilities $0, \frac14, \frac12, \frac34, 1$

* In other words: minimum, _first quartile_, median, _third quartile_, maximum

```{r}
lapply(s, quantile, probs = c(0, 0.25, 0.5, 0.75, 1),  na.rm = TRUE)
```

---

* We usually compare these quantiles _graphically_ using a __box and whisker plot__

```{r}
bwplot(Gender ~ Height, data = NHANES, subset = Age >= 20, coef = 0)
```

---

* The _compactness_ of this design makes it easier to compare multiple groups together

```{r}
bwplot(Race1 ~ Height | Gender, data = NHANES, subset = Age >= 20, coef = 0)
```

---

* Common axes makes comparison easier

```{r}
bwplot(Height ~ Gender | Race1, data = NHANES, subset = Age >= 20, layout = c(5, 1), coef = 0)
```

---

* Similar plot for a different variable: `Weight`

```{r}
bwplot(Weight ~ Gender | Race1, data = NHANES, subset = Age >= 20, layout = c(5, 1), coef = 0)
```

---

* What happens if we drop the `coef = 0`? Plots "outlier" data points that are too extreme (if Normal)

```{r}
bwplot(Weight ~ Gender | Race1, data = NHANES, subset = Age >= 20, layout = c(5, 1))
```

---

* We can go back to Q-Q plots if we see something unusual

```{r}
qqmath( ~ Weight | Race1, data = NHANES, subset = Age >= 20, layout = c(5, 1),
       groups = Gender, auto.key = TRUE, grid = TRUE)
```


---

* Sometimes transforming the data can make the distribution "closer" to Normal

```{r}
qqmath( ~ Weight^(1/3) | Race1, data = NHANES, subset = Age >= 20, layout = c(5, 1),
       groups = Gender, auto.key = TRUE, grid = TRUE)
```

---

* Why do we care about the Normal distribution?

--

* Mainly because

	* It is good approximation for many observed distributions
	
	* It is defined by just __two__ parameters: the median and the length of the box
	
	* The length of the box is known as the __IQR__ or _inter-quartile range_

--

* Many statistical problems have "nice" solutions when the underlying population is Normal

* So one focus of initial "data exploration" is to decide if observed data look like Normal

--

* This means that we need to be able to tell when data are _not_ like Normal

* This is possible from a Q-Q plot, but it is not as easy to describe the __type of departure__ 

--

* For this, another tool known as the __histogram__ is more useful

---

layout: true

# Data binning and histograms

---

* The idea of binning is useful in a variety of contexts

* Motivation from a data summary perspective:

	* Categorical or discrete data can be summarized by table counts (without loss of information)

	* Binning provides a similar summary for continuous data (with some loss of information)

--

* Another motivation is probabilistic, which we will discuss later

---

* Consider the NHANES height data (after removing `NA`-s and rows with Age less than 20)

```{r}
str(nhsub2$Height)                          # raw data ~ 7000 records
T1 <- table(nhsub2$Height); str(T1)         # table of unique values (lossless summary) ~ 500 values
```

--

```{r}
T2 <- table(round(nhsub2$Height)); str(T2)  # table of rounded values (summary with some loss) ~ 60 values
```

---

```{r}
xyplot(as.numeric(T1) ~ as.numeric(names(T1)), type = "h")
```

---

```{r}
xyplot(as.numeric(T2) ~ as.numeric(names(T2)), type = "h")
```

---

```{r}
xyplot(as.numeric(T2) ~ as.numeric(names(T2)), type = "h",
       panel = panel.barchart, horizontal = FALSE, origin = 0)
```

---

```{r}
T3 <- table(10 * round(nhsub2$Height / 10)); # round to 10th digit
xyplot(as.numeric(T3) ~ as.numeric(names(T3)), type = "h",
       panel = panel.barchart, box.width = 8, horizontal = FALSE, origin = 0)
```

---

* Histograms basically generalize this kind of rounding

* They are determined by a sequence of 'break points' defining bins (covering all the data)
	
* The data are summarized by the number (or relative frequency) of data points falling in each bin

--

* Data points may coincide with bin boundaries (break points)

* We need to be careful to specify which bin such data points should be counted in

--

* Histogram bins all _usually_ have the same width (the break points are equally spaced)

---

* Counting the number of data points in each bin is not trivial

* R has three related functions for this:

	* `cut()` converts numeric data into categories defined by breakpoints
	
	* `findInterval()` is similar but more efficient with a different return value
	
	* `hist()` computes the bin _counts_ directly
	
---

```{r}
cut(nhsub2$Height, breaks = seq(125, 205, by = 10)) |> head(30)
```

--

```{r}
cut(nhsub2$Height, breaks = seq(125, 205, by = 10)) |> table()
```

---

```{r}
hist(nhsub2$Height, breaks = seq(125, 205, by = 10), plot = FALSE) |> str()
```

---

```{r}
histogram(~ Height, NHANES, breaks = seq(125, 205, by = 10), subset = Age >= 20, type = "count")
```

---

```{r}
histogram(~ Height, NHANES, breaks = seq(125, 205, by = 1), subset = Age >= 20, type = "percent")
```

---

```{r}
histogram(~ Height, NHANES, nint = 30, subset = Age >= 20) # specify number of bins
```

---

```{r}
histogram(~ Height | Gender, NHANES, nint = 30, subset = Age >= 20, layout = c(1, 2))
```

---

layout: false

# Histograms vs bar charts

* In some ways similar, but also different

* Barcharts are meant for discrete or categorical data (with pre-specified categories)

* Histograms are meant for continuous data

--

* This is reflected in the lack of _gaps_ between bars in a histogram

--

* Relation with probability theory

	* Bar charts estimates the population _probability mass function_ (categorical / discrete)
	
	* Histograms estimate the population _probability density function_ (continuous)
	
* We will discuss these theoretical connections later

---

# Counting bin frequencies

* Easy with computers

* Not so easy without one

--

* Example: height data from survey

```{r}
survey <- read.csv("https://deepayan.github.io/BSDS/2024-01-DE/data/bsds-survey.csv")
height <- round(survey$height)
height
```

---

# Stem and leaf plot

* A very crude way of counting

```{r}
stem(height)
```

---

layout: true

# Histograms for Normal data

---

```{r}
c(histogram(~ rnorm(  10)), histogram(~ rnorm( 10)), histogram(~ rnorm(  10))) |>
    update(grid = TRUE, aspect = 1)
```

---

```{r}
c(histogram(~ rnorm(  50)), histogram(~ rnorm( 50)), histogram(~ rnorm(  50))) |>
    update(grid = TRUE, aspect = 1)
```

---

```{r}
c(histogram(~ rnorm( 100)), histogram(~ rnorm( 100)), histogram(~ rnorm( 100))) |>
    update(grid = TRUE, aspect = 1)
```

---

```{r}
c(histogram(~ rnorm( 500)), histogram(~ rnorm( 500)), histogram(~ rnorm( 500))) |>
    update(grid = TRUE, aspect = 1)
```

---

```{r}
c(histogram(~ rnorm(5000)), histogram(~ rnorm(5000)), histogram(~ rnorm(5000))) |>
    update(grid = TRUE, aspect = 1)
```

---

layout: false
class: center, middle

# Questions?





