---
layout: remark
title: Efficiency and Robustness of Estimators
subtitle: Statistics I — Data Exploration
author: Deepayan Sarkar
mathjax: true
---


```{r opts, echo = FALSE, results = "hide", warning = FALSE, message = FALSE}
opts_chunk$set(cache = TRUE, cache.path='~/knitr-cache/de-robustness/', autodep = TRUE,
               comment = "", warning = TRUE, message = TRUE,
			   ## engine.path = list(python = "/usr/bin/python3"),
               fig.width = 12, fig.height = 5,
               dev = "svglite", dev.args = list(pointsize = 12),
               knitr.table.format = "html",
			   fig.path='figures/robustness-')
options(warnPartialMatchDollar = FALSE, width = 80)
suppressWarnings(require(lattice))
suppressWarnings(require(latticeExtra))
```



<div>
$$
\newcommand{\sub}{_}
\newcommand{\set}[1]{\left\lbrace {#1} \right\rbrace}
\newcommand{\nseq}[1]{ {#1}\sub{1}, {#1}\sub{2}, \dotsc, {#1}\sub{n} }
$$
</div>


---

layout: true

# Motivation

---

* Summary statistics like sample mean and sample median are called _estimators_

* They "estimate" corresponding quantities of the _population_ (known as _parameters_)

--

* Why are we interested in _population mean_ or _population median_?

	* Because they minimize corresponding loss functions for the population
	
	* As a result, they can serve as measures of location for the population
	
	* Can be compared for sub-populations
	
---

* We typically consider symmetric populations (otherwise we try to transform)

* A population or distribution $X$ is symmetric about $a$ if 

$$
P(X \leq a - q) = P(X \geq a + q) \text{ for all } q \in \mathbb{R}
$$

--

* Important property of symmetric distributions: _mean and median are the same_

	* If $X$ is symmetric about $a$, then $a$ is both mean and median
	
	* This $a$ is therefore a natural measure of _location_

--

* We usually summarize such distributions by this location parameter + a scale parameter
	
* For now, we will only consider statistics that are estimators of
  location or spread

---

layout: true

# Desirable properties of estimators

---

* Consistency

	* As sample size increases, sample statistic (estimator) should "converge" to parameter
	
	* This is basic requirement without which estimators are not useful
	
--

* Efficiency

	* Given two consistent estimators, the one that converges more quickly is more _efficient_

--

* Robustness

	* Given two consistent estimators, the one that is less affected
      by perturbations is more _robust_

---

* Consistency, Efficiency, and Robustness are _theoretical_ properties

* We do not yet know enough probability theory to analyze them theoretically

--

* However, these are easy to study using simulation

--

* What do we need for a simulation study?

	* A distribution or "model" representing our population

	* Define parameters of interest
	
	* Estimators that we want to compare
	
	* Decide what types of perturbations we want to consider

---

layout: true

# Model 1: The Uniform distribution

---

* Recall exercise: Consider a very large fixed number $M = 10^6 =
  1000000$. Fix $n = 5001$.

* Draw a _random sample_ $\nseq{X}$ from the discrete uniform
  distribution on $\set{0, 1, 2, \dotsc, M}$

* Define $\nseq{Y}$ as $Y\sub{i} = \frac{X\sub{i}}{M}$

* Define 

	* $M\sub{1} = \frac1n \sum\limits\sub{i=1}^n Y\sub{i}$ (the sample mean)

	* $M\sub{2} = Y\sub{(2501)}$ (the sample median)

---

* Let us simulate this

```{r}
M <- 10^6; n <- 5001
m1 <- replicate(10000, mean(sample(M, n, replace = TRUE) / M))
m2 <- replicate(10000, median(sample(M, n, replace = TRUE) / M))
```

--

```{r,fig.height=3}
bwplot(which ~ data, make.groups(mean = m1, median = m2))
```

---

* From now on, we will use a _continuous_ version of the Uniform distribution

```{r}
n <- 5001
m1 <- replicate(10000, mean(runif(n)))
m2 <- replicate(10000, median(runif(n)))
```

```{r,fig.height=3}
bwplot(which ~ data, make.groups(mean = m1, median = m2))
```

---

* Summary

	* The theoretical location parameter (mean / median) is $0.5$

	* Both mean and median from a sample of $n = 5001$ are generally close

	* The mean seems to be _closer_ overall (suggesting higher _efficiency_)

---

* Distribution of mean and median

```{r}
histogram( ~ data | which, make.groups(mean = m1, median = m2), nint = 35)
```


---

layout: true

# Model: The Normal distribution $N(\mu, \sigma^2)$

---

* As mentioned earlier, the Normal distribution is a very commonly used model

* It is continuous distribution that approximates many real life distributions 

	* Including the distributions of $M\sub{1}$ and $M\sub{2}$ above

--

* It is a symmetric distribution

* It is defined by two parameters: location $\mu$ and scale $\sigma$

---

* Can be simulated in R using `rnorm(n, mean = mu, sd = sigma)`

```{r}
histogram( ~ rnorm(10000, mean = 0.5, sd = 0.01), nint = 35)
```




---

* Repeat the experiment above with Normal, and a smaller sample size

```{r}
n <- 50
m1 <- replicate(10000, mean(rnorm(n, mean = 5, sd = 3)))
m2 <- replicate(10000, median(rnorm(n, mean = 5, sd = 3)))
```

```{r,fig.height=3}
bwplot(which ~ data, make.groups(mean = m1, median = m2))
```


---

* Increase sample size (to check consistency)

```{r}
n <- 500
m1 <- replicate(10000, mean(rnorm(n, mean = 5, sd = 3)))
m2 <- replicate(10000, median(rnorm(n, mean = 5, sd = 3)))
```

```{r,fig.height=3}
bwplot(which ~ data, make.groups(mean = m1, median = m2))
```

---

* Summary

	* The theoretical location parameter (mean / median) is $5$

	* Both sample mean and sample median are close to this value, for both sample sizes
	
	* They are closer for larger sample size (suggesting both estimators are _consistent_)

	* The sample mean again seems to be _closer_ (suggesting higher _efficiency_)


---

layout: false

# Robust estimation of location

* How do these behave when data is "contaminated"?

* We know that

    * Mean can be changed by an arbitrary amount by changing a single observation

    * Median can be changed arbitrarily only by changing more than 50% observations

    * This comes at the cost that median is less "efficient" for many distributions

---

layout: true

# Other estimators

---

* Another estimator: _trimmed mean_ or _truncated mean_

* Definition:

	> The $\alpha$-trimmed mean of a vector of observations
	> $\boldsymbol{X} = (\nseq{X})$ is the mean of the observations
	> that remain after discarding a proportion $\alpha$ of the
	> observations from both ends.

--

* Can be computed in R using the `mean()` function, with argument `trim` specifying $\alpha$

```{r}
mean(runif(30), trim = 0.1) # 10% trimming on both ends
```

---

* Another estimator: _winsorized mean_

* Definition:

	> the mean after _replacing_ (not discarding) a proportion
	> $\alpha$ of the most extreme observations from both ends by the
	> most extreme of the remaining values
	
* The process of transforming a vector $\boldsymbol{X}$ in this way is known as _winsorizing_
	
--

* No built-in function in R, but easy to implement

```{r}
winsorize <- function(x, trim = 0.25)
{
    q <- quantile(x, c(trim, 1-trim))
    x[x < q[1]] <- q[1]
    x[x > q[2]] <- q[2]
    x
}
winsorized.mean <- function(x, trim = 0.25) mean(winsorize(x, trim = trim))
```

--

```{r}
winsorized.mean(runif(30), trim = 0.1) # 10% trimming on both ends
```

---

* Both these estimators are clearly designed to protect against extreme outliers

* How do they compare in terms of robustness and efficiency

* The easiest way to understand this is through simulation.


---

layout: true

# Relative efficiency

---

* Consider two estimators $T\sub{1}$ and $T\sub{2}$

* Define the relative efficiency of $T\sub{1}$ w.r.t. $T_2$ (for a given underlying distribution) as

$$
RE(T\sub{1}; T_2) = \frac{\text{MSE}(T_2)}{\text{MSE}(T\sub{1})}
$$

--

* $\text{MSE}(T)$ stands for Mean Squared Error, and can be estimated using average squared-error loss

--

* What are relative efficiencies of median w.r.t. mean? What about other estimators?

* Trying to obtain MSE theoretically is often difficult
  
* But we could use simulation to get a rough idea

---

layout: true

# Relative efficiency in the Normal population

---

* Extend previous example to study efficiency

* Helper functions

```{r}
trimmed.mean <- function(x, trim = 0.25)
{
    mean(x, trim = trim)
}
simulate_estimator <- function(estimator, rfun, n, NREP = 10000)
{
    replicate(NREP, estimator(rfun(n)))
}
mse <- function(T, theta)
{
    mean((T - theta)^2)
}
```

---

* Simulate with $n = 10$

```{r}
rdist <- function(n) rnorm(n, mean = 5, sd = 3) # Normal(5, 3^2)
```

```{r}
sim10 <- list(mean = simulate_estimator(mean, rdist, n = 10),
              median = simulate_estimator(median, rdist, n = 10),
              tmean = simulate_estimator(trimmed.mean, rdist, n = 10),
              wmean = simulate_estimator(winsorized.mean, rdist, n = 10))
```

---

```{r}
do.call(make.groups, sim10) |> bwplot(which ~ data)
```

---

* Relative Efficiency w.r.t. mean

```{r}
with(sim10, round(100 * mse(mean, 5) / mse(median, 5))) # median
with(sim10, round(100 * mse(mean, 5) / mse(tmean, 5)))  # trimmed mean
with(sim10, round(100 * mse(mean, 5) / mse(wmean, 5)))  # winsorized mean
```

---

* Asymptotic relative efficiency $ARE(T\sub{1}; T_2)$ 

	* Limiting value of relative efficiency as $n \to \infty$

	* We can approximate this by choosing $n$ to be large (here $n = 5000$)

```{r}
rdist <- function(n) rnorm(n, mean = 5, sd = 3) # Normal(5, 3^2)
```

```{r}
sim5000 <- list(mean = simulate_estimator(mean, rdist, n = 5000),
                median = simulate_estimator(median, rdist, n = 5000),
                tmean = simulate_estimator(trimmed.mean, rdist, n = 5000),
                wmean = simulate_estimator(winsorized.mean, rdist, n = 5000))
```

---

```{r}
do.call(make.groups, sim5000) |> bwplot(which ~ data)
```

---

* (Asymptotic) Relative Efficiency w.r.t. mean

```{r}
with(sim5000, round(100 * mse(mean, 5) / mse(median, 5))) # median
with(sim5000, round(100 * mse(mean, 5) / mse(tmean, 5)))  # trimmed mean
with(sim5000, round(100 * mse(mean, 5) / mse(wmean, 5)))  # winsorized mean
```


* For comparison, the exact $ARE$ of the median is $\frac{2}{\pi} = 63.6\%$ <!-- for pseudo-proof, note that $\Phi(median) \sim Beta(n/2, n/2)$ for $N(0,1)$  -->

* This is when the data comes from a Normal distribution


---

layout: true

# Relative efficiency for distributions with heavier tails

---

* Consider the $t$ distribution with parameter `df = 5`

* Superficially similar no Normal (symmetric with single peak)

* However, tails are heavier than Normal (extreme observations more likely)

```{r}
rdist <- function(n) 5 + 3 * rt(n, df = 5)
```

```{r,fig.height=3.5}
histogram(~ rdist(10000), nint = 35)
```

	
---

```{r}
simt5 <- list(mean = simulate_estimator(mean, rdist, n = 5000),
              median = simulate_estimator(median, rdist, n = 5000),
              tmean = simulate_estimator(trimmed.mean, rdist, n = 5000),
              wmean = simulate_estimator(winsorized.mean, rdist, n = 5000))
```

```{r,fig.height=3}
do.call(make.groups, simt5) |> bwplot(which ~ data)
```

---

* (Asymptotic) Relative Efficiency w.r.t. mean

```{r}
with(simt5, round(100 * mse(mean, 5) / mse(median, 5))) # median
with(simt5, round(100 * mse(mean, 5) / mse(tmean, 5)))  # trimmed mean
with(simt5, round(100 * mse(mean, 5) / mse(wmean, 5)))  # winsorized mean
```

---

layout: true

# Relative efficiency for contaminated Normal

---

* Suppose data is a mixture of $N(170, 10^2)$ with probability
  $1-\epsilon$ and $N(80, 30\sigma^2)$ with probability $\epsilon$

```{r}
rdist <- function(n)
{
    x1 <- rnorm(n, mean = 170, sd = 10)
    x2 <- rnorm(n, mean = 80, sd = 30)
    ifelse(runif(n) < 0.99, x1, x2) # 1% contamination
}
```


```{r,fig.height=3.5}
histogram(~ rdist(10000), nint = 35)
```

	
---

```{r}
simmix <- list(mean = simulate_estimator(mean, rdist, n = 5000),
               median = simulate_estimator(median, rdist, n = 5000),
               tmean = simulate_estimator(trimmed.mean, rdist, n = 5000),
               wmean = simulate_estimator(winsorized.mean, rdist, n = 5000))
```

```{r,fig.height=3}
do.call(make.groups, simmix) |> bwplot(which ~ data)
```

---

* (Asymptotic) Relative Efficiency w.r.t. mean

```{r}
with(simmix, round(100 * mse(mean, 170) / mse(median, 170))) # median
with(simmix, round(100 * mse(mean, 170) / mse(tmean, 170)))  # trimmed mean
with(simmix, round(100 * mse(mean, 170) / mse(wmean, 170)))  # winsorized mean
```

---

* Exercise: Try variations of this

	* Efficiency as function of trimming parameter $\alpha$
	
	* Efficiency as function of mixture proportion $\epsilon$
	
	* Efficiency as function of mixture location


<!--

Possible modifications: 

- Compute MSE instead of var(), need to provide true value of estimate

- Draw plot of variances as function of t-df, or epsilon above

- Can we do this theoretically?

	- Assume Uniform median U ~ Beta(n/2, n/2) is asymp normal with same mean / var
	- Compute asymp dist of F^{-1}(U) using Delta method
	- Compute MSE for the mixture model - is that feasible?
	- Plot as function of contamination variance for different epsilon
	- Is median better even for contamination variance lower than the actual variance?

-->


---

layout: true

# Sensitivity / influence function

---

* Natural question: How much does changing one observation change the estimate $T$?

* This is measured by the _empirical influence function_ or _sensitivity curve_

$$
SC(x; x\sub{1}, \dotsc, x\sub{n-1}, T) = \frac{ T(x\sub{1}, \dotsc, x\sub{n-1}, x) - T(x\sub{1}, \dotsc, x\sub{n-1}) }{ 1/n } 
$$

* We are usually interested in limiting behaviour as $n \to \infty$

<!-- 

* The population analogue, independent of the sample $x\sub{1}, \dotsc,
  x\sub{n-1}$, is known as the _influence function_

$$
IF(x; F, T) = \lim\sub{\epsilon \to 0} \frac{ T( (1-\epsilon) F + \epsilon \delta_x) - T(F) }{\epsilon}
$$

* Here $\delta_x$ is a point mass at $x$

-->

---

* How does `mean(c(x, xnew))` change as function of `xnew`?

* How do other estimates change?

```{r}
nn <- 50
x <- rnorm(nn, mean = 0, sd = 1)
fivenum(x)
xx <- seq(-5, 5, 0.01)
sensitivity <-
    data.frame(xx = xx,
               mean = nn * (sapply(xx, function(xnew) mean(c(x, xnew))) - mean(x)),
               median = nn * (sapply(xx, function(xnew) median(c(x, xnew))) - median(x)),
               tmean = nn * (sapply(xx, function(xnew) trimmed.mean(c(x, xnew))) - trimmed.mean(x)),
               wmean = nn * (sapply(xx, function(xnew) winsorized.mean(c(x, xnew))) - winsorized.mean(x)))
```

---

```{r}
xyplot(mean + median + tmean + wmean ~ xx, sensitivity, type = "l", outer = TRUE,
       xlab = "Additional observation", ylab = "Sensitivity", grid = TRUE) +
    latticeExtra::layer(panel.rug(x = .GlobalEnv$x))
```

---

layout: false

# Important measure of Robustness: Breakdown Point

* Defined as the proportion of the sample size that must be perturbed
  to make the estimate unbounded

* 50% is the best we can hope for

* For location, 

	* mean has 0% breakdown point (one out of $n$), 
	
	* median has 50%,
	
	* others depend on amount of trimming
	
	


---

# Estimators of scale

* We can similarly consider estimators of scale $\sigma$

* Common estimators:

    * Sample standard deviation (`sd` in R)

    * Mean absolute deviation from median

    * Median absolute deviation (MAD) from median (`mad` in R)

    * Inter-quartile range (`IQR` in R)

* May need scaling for normal distribution:

```{r}
T1 <- sd
T2 <- function(x, ...) mean(abs(x - mean(x))) / sqrt(2/pi)
T3 <- function(x, ...) median(abs(x - median(x))) / sqrt(qchisq(0.5, df = 1))
T4 <- function(x, ...) IQR(x) / diff(qnorm(c(0.25, 0.75)))
```

--

* Exercise: Estimate relative efficiency for the models considered earlier


---

layout: false
class: center, middle

# Questions?





