---
layout: remark
title: A Sample Session in R
subtitle: Data Analysis with R and Python
author: Deepayan Sarkar
mathjax: true
---

```{r opts, echo = FALSE, results = "hide", warning = FALSE, message = FALSE}
opts_chunk$set(cache = TRUE, cache.path='~/knitr-cache/darp-r-session/', autodep = TRUE,
               comment = "", warning = TRUE, message = TRUE,
			   ## engine.path = list(python = "/usr/bin/python3"),
               fig.width = 12, fig.height = 7,
               dev = "svglite", dev.args = list(pointsize = 20),
               knitr.table.format = "html",
			   fig.path='figures/1-rsession-')
options(warnPartialMatchDollar = FALSE, width = 80)
```


# Starting and Interacting with R

* R is typically used interactively

* When we start R, the command window (or console) displays a prompt, typically `>`

* We use R by entering an expression to be evaluated

* R evaluates the expression and prints the result

```{r}
1 + 2
```

* It then provides a new prompt and waits for more input


<div>
$$
\newcommand{\sub}{_}
$$
</div>



---

# Infix and Prefix Notation

* R uses **infix** notation for standard arithmetic operations, e.g., 

```r
1 + 2
```

--

* The corresponding **prefix** notation would look something like

```
+ 1 2
```

--

* This is actually what R does internally, using _function notation_

```{r}
`+`(1, 2)
```

* In general, R expressions are typically function calls of the form `f(a, b)`

---

layout: true

# Basic principles: Data types

---

* R can handle may different kinds of data

* Basic classification: *simple data* and *compound data*

* **Simple Data** includes:

	* Numbers (numeric values, including integers and floating-point numbers):
```r
1     # an integer  
-3.14 # a floating point number
```
--

	*  Logical values:
```r
 TRUE # true  
FALSE # false (or T and F shortcuts)
```
--

	* Strings (enclosed in single or double quotes):
```r
"This is a string 1 2 3 4"
```

---

* **Compound Data** primarily consists of

	* **vectors** : ordered collections of elements of the same type
	
	* **lists** : ordered collections with elements of possibly different types

--
* We commonly define compound data using the concatenate function `c()`:
```{r}
c(1, 2, 3)
```

---

* We can also have **Symbols** which are used for naming variables or functions:

```r
x  
gdp.data  
this_is_a_symbol
```

---

layout: false

# The REPL

* An R session involves interaction between the user and the console (_listener_)

* When we enter an expression, the listener passes it to the _evaluator_

--

* Basic rule:

	* Everything is evaluated

	* The results are (usually) printed
	
	* Once done, listener goes back to listening

--

<br/>

.center[

This is known as the __Read-Eval-Print-Loop__ or __REPL__

]

---

layout: true

# Evaluation rules

---

* Numbers and strings evaluate to themselves:

```{r}
10
"Hello"  
```

---

* Expressions can involve _functions_ 

```{r}
sqrt(10)
nchar("Hello")
```

--

* Functions are applied using parentheses

--

* Function name precedes arguments, mirroring **prefix** notation

	* For example ` min(x, y) ` instead of ` min x y `

	* Parentheses optional (if number of arguments are known) but helpful
	
	* Compare ` min(pi, sqrt(10)) ` and ` min pi sqrt 10 `

	* Essential when number of arguments of function unknown

---

* Evaluation can be **suppressed** by _quoting_ an expression

```{r}
quote(min(pi, sqrt(10)))
```

--

* This turns out to be a very interesting feature that we will return to later

* Python does **not** have a similar feature (R and Python are otherwise very similar)

