---
layout: remark
title: Basic Usage of R
subtitle: Data Analysis with R and Python
author: Deepayan Sarkar
mathjax: true
---

	
# R is a full programming language

* Variables

* Functions

* Control flow structures

	* For loops, while loops

	* If-then-else (branching)

--

* Distinguishing features

	* Focus on _vectors_ and _vectorized operations_ 

	* Treatment of _functions_ at par with other object types
	
	* _Functional language_ with minimal _side effects_ of function calls


```{r opts, echo = FALSE, results = "hide", warning = FALSE, message = FALSE}
library(knitr)
library(kableExtra)
opts_chunk$set(cache = TRUE, cache.path='~/knitr-cache/darp-basics/', autodep = TRUE,
               comment = "", warning = TRUE, message = TRUE,
               fig.width = 12, fig.height = 5.5,
               dev = "svglite", dev.args = list(pointsize = 15, bg = "transparent"),
               knitr.table.format = "html",
               fig.path='figures/2-basics-')
## opts_chunk$set(engine.path = list(python = '/usr/local/bin/python3'))
options(warnPartialMatchDollar = FALSE, width = 90) # 110
```



<div>
$$
\newcommand{\sub}{_}
$$
</div>


---

# R is easily extensible

* Most standard data analysis methods are already implemented

* Can be extended by writing add-on packages

* Thousands of add-on packages are available

---

# Major concepts we need to know

* Variables (in the context of programming)

--

* Data structures needed for data analyis

--

* Functions (set of instructions for performing a procedure)

---

# Variables

* Variables are symbols that may be associated with different values

* Computations involving variables are done using their current value

```r
sqrt(x)
```

```
Error: object 'x' not found
```

---

# Variables

* Variables are symbols that may be associated with different values

* Computations involving variables are done using their current value


```{r}
x <- 10 # assignment
sqrt(x)
x <- -1
sqrt(x)
x <- -1+0i
sqrt(x)
```

---

# Data structures for data analysis

* Vectors

* Matrices 

* Data frames (a spreadsheet-like data set)

* Lists (general collection of objects)

---

# Atomic vectors

* Indexed collection of homogeneous scalars, can be 

    * Numeric / Integer / Complex

    * Character

    * Logical (`TRUE` / `FALSE`)
	
    <!-- * Categorical (factor) -->

--

* Missing values are allowed, indicated as `NA`


--

* Elements are indexed starting from 1


--

* $i$th element of vector `x` can be extracted using `x[[i]]`

* There are also more sophisticated forms of (vector) indexing



---

# Atomic vectors: examples

```{r}
month.name # built-in 
```


--

```{r}
x <- rnorm(10)
x
```


--

```{r}
x[[3]] # third element of x
```

---

# Atomic vectors: examples

```{r}
str(x) # useful function
str(month.name)
```

---

# Creating atomic vectors

* Constructor functions

```{r}
numeric(10)
logical(5)
character(5)
```

---

# Scalars are also vectors

* "Scalars" are just vectors of length 1

```{r}
str(numeric(2))
str(numeric(1))
str(0)
```

---

# Vectors can have zero length

* Vectors can have length zero 

```{r}
numeric(0)
logical(0)
length(character(0))
```

--

```{r}
length(NULL)
```

---

# Combining vectors using `c()` 

* Vectors can also be created by combining smaller vectors

* For example, vectors `x` and `y` can be combined using `c(x, y)`

```{r}
c(1:5, numeric(3))
```

--

* Any number of vectors can be combined

* This is a common way to build up a vector using scalars

```{r}
c(2, 4, 6, 9, 11)
```

---

# Combining vectors of different types

* Atomic vectors of different types cannot be combined

* Attempting to do so will convert into one of the types

```{r}
c(1:5, c(TRUE, FALSE))
c(1:5, month.name[[1]])
```

--


```{r}
c(1:5, c(TRUE, FALSE), month.name[[1]])
c(c(1:5, TRUE, FALSE), month.name[[1]])
```

---


# Example: Life Expectancy Dataset

* Life expectancy in different countries over time

```{r,echo=FALSE}
## library(pander)
g <- read.table("data/gapminder.tsv", sep = "\t", header = TRUE)
keep.countries <- c("Australia", "France", "India", "Zimbabwe")
lexp.long <- subset(g, country %in% keep.countries, select = c(country, year, lifeExp))
lexp.wide <- reshape(lexp.long, direction = "wide", idvar = "year", timevar = "country")
names(lexp.wide) <- gsub("lifeExp.", "", names(lexp.wide), fixed = TRUE)
rownames(lexp.wide) <- NULL
kable (lexp.wide)
```

---

layout: true

# Life Exepectancy in France

---

```{r}
year <- c(1952, 1957, 1962, 1967, 1972, 1977,
          1982, 1987, 1992, 1997, 2002, 2007)
year
```


--

```{r}
lexp_france <- c(67.41, 68.93, 70.51, 71.55, 72.38, 73.83, 74.89, 76.34,
              77.46, 78.64, 79.59, 80.657)
lexp_france
```


---


```{r}
year <- seq(1952, 2007, by = 5)
year
```

---

```{r lifexp-france}
plot(year, lexp_france, pch = 16)
```


---

```{r}
lexp_france[[2]] - lexp_france[[1]]
```

---

```{r}
c(lexp_france[[2]] - lexp_france[[1]], lexp_france[[3]] - lexp_france[[2]],
  lexp_france[[4]] - lexp_france[[3]], lexp_france[[5]] - lexp_france[[4]],
  lexp_france[[6]] - lexp_france[[5]], lexp_france[[7]] - lexp_france[[6]],
  lexp_france[[8]] - lexp_france[[7]], lexp_france[[9]] - lexp_france[[8]],
  lexp_france[[10]] - lexp_france[[9]], lexp_france[[11]] - lexp_france[[10]],
  lexp_france[[12]] - lexp_france[[11]])
```

---

```{r}
d <- numeric(0)
```


--

```{r}
for (i in 1:11) {
    d <- c(d, lexp_france[[i+1]] - lexp_france[[i]])
}
d
```

---

```{r}
lexp_france[-1] - lexp_france[-12]
```


--

```{r}
diff(lexp_france)
```

---

```{r}
d <- diff(lexp_france)
median(d)
mean(d)
```

---

```{r lexp-diff-france}
plot(d, pch = 16, type = "o", ylab = "difference", xlab = "period")
```

---

layout: false

# Types of vector indexing

* Indexing refers to extracting subsets of data

* R supports several kinds of indexing:

	* Indexing with a vector of positive integers

	* Indexing with a vector of negative integers
	
	* Indexing with a logical vector

	* Indexing with a vector of names


---

# The empty index

* A vector indexing operation has the form `x[index]`

--

* The most basic form is an empty index, which selects all elements

```{r}
month.name[]
```


---

# Indexing with an integer vector

* For integer indexing, `index` is an integer vector

```{r}
month.name[c(2, 4, 6, 9, 11)]
```


--

* Elements can be repeated

```{r}
month.name[c(2, 2, 6, 4, 6, 11)]
```


---

# Indexing with an integer vector

* "Out-of-bounds" indexing gives `NA` (missing)

```{r}
month.name[13]
```


--

```{r}
seq(1, by = 2, length.out = 8)
```


--

```{r}
month.name[seq(1, by = 2, length.out = 8)]
```



---

# Indexing with an integer vector

* Indexing with a scalar (vector of length 1) also works:

```{r}
month.name[2]
```


--

* This is usually the same as `x[[index]]`

```{r}
month.name[[2]]
```


--

* However, these differ in the behaviour when an index is out of bound

```{r}
month.name[15]
month.name[[15]]
```


---

# Indexing with a vector of negative integers

* Negative integers omit the specified entries

```{r}
month.name[-2]
month.name[-c(2, 4, 6, 9, 11)]
```


--

* Cannot be mixed with positive integers

```{r}
month.name[c(2, -3)]
```



---

# Indexing with 0

* Zero has a special meaning - it doesn't select anything

```{r}
month.name[0]
month.name[integer(0)] ## same as empty index
month.name[c(1, 2, 0, 11, 12)]
month.name[-c(1, 2, 0, 11, 12)]
```


---

# Indexing with a logical vector

* Indexing by logical vector: select `TRUE` elements

```{r}
month.name[c(TRUE, FALSE, FALSE)] # index replicated
```


---

# Indexing with a logical vector

* Indexing by logical vector: select `TRUE` elements

```{r}
i <- substring(month.name, 1, 1) == "J"
i
```


--

```{r}
month.name[i]
```



---

# Indexing with a logical vector

<!-- * Common use: extract subset satisfying a certain condition (also called "filtering") -->

```{r}
(x <- rnorm(20))
x > 0
x[x > 0]
mean(x[x > 0])
```



---

# Converting a logical index vector to integer

* Logical indexing may be replaced by integer indexing using `which()`

```{r}
i
which(i)
```


--

```{r}
month.name[ which(i) ]
```


--

```{r}
month.name[ -which(i) ] # same as month.name[ !i ]
```


---

# Converting a logical index vector to integer

* But be careful about zero-length indices

```{r}
which(substring(month.name, 1, 1) == "B")
month.name[ which( substring(month.name, 1, 1) == "B") ]
-which(substring(month.name, 1, 1) == "B") 
month.name[ -which( substring(month.name, 1, 1) == "B") ]
```


---

layout: true

# Indexing with a vector of names

---

- Vectors can optionally have names — one for each element

- These are usually informative labels


--

- Example: quantiles of a Normal random sample

```{r}
x <- rnorm(100)
qx <- quantile(x)
qx
names(qx)
```


--

```{r}
names(x) # no names
```


---

- When present, names may be used to identify elements

- Indexing with names works in the same way as positive integers

- Instead of position, the corresponding named element is selected

```{r}
qx[["50%"]] ## extracting a single element using scalar indexing
qx["50%"]   ## extracting a single element with vector indexing
qx[c("25%", "75%")]  ## extracting multiple elements
```


---

* Inter-quartile range

```{r}
diff(qx[c("25%", "75%")])
```


--

```{r}
IQR(x)
```


---

- Unmatched names are treated like out-of-bound indexes

```{r}
qx[["95%"]]
qx["95%"]
```



