---
layout: remark
title: Lists and Data Frames
subtitle: Data Analysis with R and Python
author: Deepayan Sarkar
mathjax: true
---


# Lists

* Lists are vectors with arbitrary types of components

--

* Individual elements can be extracted using `x[[i]]`

* Vector indexing by `x[i]` also works in the usual way

--

* A list may or may not have names

* Lists with names have a special type of extraction operator: `$`


```{r opts, echo = FALSE, results = "hide", warning = FALSE, message = FALSE}
library(knitr)
library(kableExtra)
opts_chunk$set(cache = TRUE, cache.path='~/knitr-cache/darp-listdf/', autodep = TRUE,
               comment = "", warning = TRUE, message = TRUE,
               fig.width = 12, fig.height = 5.5,
               dev = "svglite", dev.args = list(pointsize = 15, bg = "transparent"),
               knitr.table.format = "html",
               fig.path='figures/3-listdf-')
options(warnPartialMatchDollar = FALSE, width = 90) # 110
```



<div>
$$
\newcommand{\sub}{_}
\newcommand{\nseq}[1]{ {#1}\sub{1}, {#1}\sub{2}, \dotsc, {#1}\sub{n} }
$$
</div>


---

layout: true

# Example: Results of an optimization procedure

---

* Problem: 

	* A farmer has 100 meters of fencing 
	
	* He wants to enclose a rectangular area to grow spinach
	
	* His goal is to maximize the enclosed area of the field
	
	* What dimensions (length and width) should he use?

--

* The answer is obviously that the area should be a square (with each side 25m)

--

* Easy to formulate and solve mathematically:

<div>
$$
\begin{eqnarray*}
2 * (a + b) &=& 100 \implies b = 50 - a \cr
\text{To maximize } a * b &=& a * (50 - a) = 50a - a^2 \iff \text{to minimize } a^2 - 50a \cr
\text{Differentiate and solve } &:& 2a - 50 = 0
\end{eqnarray*}
$$
</div>

---

* Suppose we wanted to use numerical optimization instead

```{r}
L <- function(a) a^2 - 50 * a
res <- optimize(L, interval = c(0, 100))
str(res)
```

---

layout: false

# Printing lists

```{r}
res
```

---

# Scalar indexing

* Standard indexing using `x[[index]]`

```{r}
res[[1]]
res[["objective"]]
```

--

* Extracting element by name using `x$name`

```{r}
res$minimum
```

---

# Vector indexing

```{r}
res[1]
res["objective"]
```

--

```{r}
res[c(2, 1, 2)]
```


---

layout: false

# Lists as containers

* This is an important use of lists

	* Represent result of an analysis

	* Collects various relevant quantities 
	
	* Result contains all quantities in single object

---

# Other algorithms for "root finding"

* Exercise: Can we write our own optimization method?

--

* Simple algorithms (root finding)

	* Grid search
	
	* Bisection method
	
	* Regula falsi

---

# Example: Confidence interval for probability $p$

* Given data: $\nseq{X}$ i.i.d. sample from Bernoulli($p$)

* We want to estimate unknown success probability $p$

--

* Natural estimator is the sample proportion $\hat{p} = \bar{X}$

--

* But $\hat{p}$ will usually not be exactly equal to $p$

* What values of $p$ are _plausible_? (Consistent with observed data)

---

# Recall: distribution of sample proportion

.scrollable500[

```{r prop-rep,fig.height = 6}
p <- 0.65
plot(1:100, type = "n", ylim = c(0, 1), ylab = "Sample proportion")
for (i in 1:50) {
    z <- rbinom(100, size = 1, prob = p)
    lines(1:100, cumsum(z) / 1:100, col = sample(colors(), 1))
}
```

]

---

layout: true

# Example: Does a random quadratic equation have real roots?

---

* Suppose $A, B, C \sim N(0, 1)$

* Are the roots of $A x^2 + B x + C = 0$ real?


--

```{r}
z <- logical(500)
for (i in 1:500) {
    A <- rnorm(1); B <- rnorm(1); C <- rnorm(1)
    D <- B^2 - 4 * A * C
    z[i] <- D >= 0
}
str(z)
```

---

```{r real-roots}
plot(1:500, cumsum(z) / 1:500, type = "l")
```

---

* One method implemented in R function `prop.test()`

```{r}
tt500 <- prop.test(sum(z), 500)
str(tt500, give.attr = FALSE)
```

--

```{r}
tt500$conf.int
```

---

* What if we only had 100 replications?

```{r}
tt100 <- prop.test(sum(z[1:100]), 100)
str(tt100, give.attr = FALSE)
```

```{r}
tt100$conf.int
```


---

layout: false

# Lists as containers

```{r}
tt500
```

---

# Why doesn't output look like a list?

* This is an important feature of R

* Uses concepts such as attributes and class

* We will discuss these later

---

layout: false

# Data Frames

* R analog of a spreadsheet

---

# Example

.scrollable500[

```{r,echo=FALSE}
kable(MASS::Cars93, "html") |> kable_styling("striped")
```

]

---

layout: true

# Data Frames

---

* R analog of a spreadsheet

* Rectangular (matrix-like) structure

--

* Each column is (usually) an atomic vector

* Different columns can be of different types

--

* Every column must have the same length

* Every column must have a name

---

* Most built-in data sets in R are data frames


.scrollable500[

```{r}
data(Cars93, package = "MASS")
Cars93
```

]


---

* Data frames are internally stored as lists (with constraints)

.scrollable500[

```{r}
str(Cars93)
```

]

---

* List-like behaviour: Columns can be extracted like a list
	
```{r}
Cars93$MPG.city
```

* Vector indexing extracts multiple columns

```{r}
head(Cars93[c(1, 4, 7)])
```

---


```{r}
carsub <- Cars93[c("Make", "MPG.city", "Weight", 
                   "Length", "EngineSize", "Man.trans.avail")]
str(carsub)
```



---

* Two-dimensional indexing

```{r}
carsub[1:6, ]
carsub[1:6, c(1, 4, 6)]
```

---

* Two-dimensional indexing

```{r}
nrow(carsub)
carsub[sample(nrow(carsub), 6), ]
```

---

* Two-dimensional indexing

```{r}
carsub[sample(nrow(carsub), 6), c("MPG.city", "Weight", "Length")]
```

---

layout: true

# Data import

---

* Statistical data are usually structured like a spreadsheet (Excel / CSV)

--

* Typical approach: read data from spreadsheet file into data frame

--

* Easiest route:

    * R itself cannot read Excel files directly

    * Save as CSV file from Excel / LibreOffice / Google Sheets

    * Read with `read.csv()` or `read.table()` (more flexible)

--

* Alternative: Use "Import Dataset" tool in RStudio


--

* Data frames can be exported as a spreadsheet file using
  `write.csv()` or `write.table()`

---

layout: false

# Exporting data as CSV file


```{r}
str(carsub)
write.csv(carsub, file = "cars.csv")
```

---

# Other data import / export utilities

* These are the most basic data input / output functions

* There are many other other specialized functions

* Low-level utilties: `scan()`, `readLines()`, `readChar()`, `readBin()`

* Various packages provide import / export to formats used by other software

* R has its own "serialization" format using `save()` and `load()`

---

layout: false

# Example: air quality data

.scrollable500[

```{r}
aqi <- read.csv("https://deepayan.github.io/BSDS/2026-01-DARP/slides/data/rkpuram-aqi.csv")
aqi
```

]













