---
layout: remark
title: Functions, Attributes, and Class
subtitle: Data Analysis with R and Python
author: Deepayan Sarkar
mathjax: true
---


# Functions

* Most useful things in R happen by calling functions

* Functions have one or more arguments

    * All arguments have names (available as variables inside the function)

	* Arguments may be compulsory or optional

	* Optional arguments usually have "default" values 

--

* Arguments may or may not be named when the function is called
	
	* Unnamed arguments are matched by position

	* Optional arguments are usually named
	
* Functions normally also have a useful "return" value


```{r opts, echo = FALSE, results = "hide", warning = FALSE, message = FALSE}
library(knitr)
library(kableExtra)
opts_chunk$set(cache = TRUE, cache.path='~/knitr-cache/darp-funcattr/', autodep = TRUE,
               comment = "", warning = TRUE, message = TRUE,
               fig.width = 12, fig.height = 5.5,
               dev = "svglite", dev.args = list(pointsize = 15, bg = "transparent"),
               knitr.table.format = "html",
               fig.path='figures/4-funcattr-')
options(warnPartialMatchDollar = FALSE, width = 90) # 110
```


<div>
$$
\newcommand{\sub}{_}
\newcommand{\nseq}[1]{ {#1}\sub{1}, {#1}\sub{2}, \dotsc, {#1}\sub{n} }
$$
</div>


---

layout: true

# Example: Delhi Air Quality data

---

```{r}
aqi <- read.csv("https://deepayan.github.io/BSDS/2026-01-DARP/slides/data/rkpuram-aqi.csv")
str(aqi)
```

---

* How is AQI (PM10) related to `no2` (Nitrogen dioxide)?

--

* Basic tools we are familiar with

	* Scatter plot
	
	* Linear Regression

---

```{r plot-pm10-no2}
plot(pm10 ~ no2, data = aqi)
```

---

```{r plot-pm10-no2-sqrt}
plot(sqrt(pm10) ~ sqrt(no2), data = aqi)
```

---

* Formula for linear regression line

<div>
$$
\begin{eqnarray*}
\hat{b} &=& \frac{\sum (x\sub{i} - \bar{x}) (y\sub{i} - \bar{y})}{\sum (x\sub{i} - \bar{x})^2} \cr
\hat{a} &=& \bar{y} - \hat{b} \bar{x}
\end{eqnarray*}
$$
</div>
	
---

* Calculation of linear regression line

```{r}
x <- sqrt(aqi$no2)
y <- sqrt(aqi$pm10)
xbar <- mean(x)
ybar <- mean(y)
sxy <- sum( (x - xbar) * (y - ybar) )
sxx <- sum( (x - xbar)^2 )
```

--

```{r}
c(xbar, ybar, sxx, sxy)
```

---

* Regression coefficients removing missing values

```{r}
x <- sqrt(aqi$no2)
y <- sqrt(aqi$pm10)
xbar <- mean(x, na.rm = TRUE)
ybar <- mean(y, na.rm = TRUE)
sxy <- sum( (x - xbar) * (y - ybar), na.rm = TRUE)
sxx <- sum( (x - xbar)^2, na.rm = TRUE)
```

```{r}
c(xbar, ybar, sxx, sxy)
```

---

* Even this may not be correct: should remove both if only one missing!

```{r}
x <- sqrt(aqi$no2)
y <- sqrt(aqi$pm10)
ok <- is.finite(x) & is.finite(y)
xbar <- mean(x[ok]); ybar <- mean(y[ok])
sxy <- sum( (x[ok] - xbar) * (y[ok] - ybar))
sxx <- sum( (x[ok] - xbar)^2)
```

```{r}
c(xbar, ybar, sxx, sxy)
```

---

* Coefficients of linear regression line

```{r}
bhat <- sxy / sxx
ahat <- ybar - bhat * xbar
c(intercept = ahat, slope = bhat)
```

--

* Standard R function to do this

```{r}
lm(sqrt(pm10) ~ sqrt(no2), data = aqi)
```

---

* Add regression line to scatter plot

```{r plot-pm10-no2-with-regline}
plot(sqrt(pm10) ~ sqrt(no2), data = aqi)
abline(ahat, bhat, col = 2)
```


---

layout: true

# Functions

---

* We have used several functions above

	* `sqrt()`, `mean()`, `sum()` — basic mathematical / summary functions
	
	* `plot()` — "high level" plotting function
	
	* `abline()` — "low level" plotting function
	
	* `lm()` — "high level" modeling function

---

* We will discuss graphics functions in more detail later

--

* `lm()` is a good example to study the behaviour of modeling functions in R

--

* Recall:

	* Functions have one or more arguments

    * All arguments have names (available as variables inside the function)

	* Arguments may be compulsory or optional

	* Optional arguments usually have "default" values 



---

* `lm()` fits a more general class of models known as _linear models_

```{r}
str(lm)
```

--

* These calls are equivalent

```{r}
fm1 <- lm(pm10 ~ no2, aqi, (no2 < 100))
fm2 <- lm(pm10 ~ no2, data = aqi, method = "qr", subset = (no2 < 100))
coef(fm1)
coef(fm2)
```

---

* Rule: named arguments are matched by name, remaining by position

* Convention: 

	- First few "standard" arguments are usually unnamed (matched by position)
	
	- Usually unnamed arguments are _not_ used after named arguments

--

* The following call is equivalent to previous two, but not recommended

```{r}
fm3 <- lm(pm10 ~ no2, data = aqi, no2 < 100)
coef(fm3)
```

---

* The return value of `lm()` is a list

.scrollable400[

```{r}
str(fm2)
```

]

---

* Individual elements can be extracted using list indexing

```{r}
fm2$coefficients
fm2$residuals
```

---

layout: true

# Attributes

---

* The names (derived from row names of the data) can be used as index 

```{r}
fm2$residuals["25"]
```

* The names associated with a vector can be obtained using `names()`

```{r}
names(fm2$residuals)
```

---

* These names are actually stored as an _attribute_ called "names"

```{r}
attr(fm2$residuals, "names")
```

---

* This is true for all vector objects, including lists

```{r}
attr(fm2, "names")
```

---

* Attributes are arbitrary R objects that can be attached to any other object

* Typically used for programming convenience, normally not seen by users

* However, some attributes are "special"

---

* The "names" attribute can be extracted using the function `names()`

* `dimnames()` similarly gives row / column names for matrices and arrays

```{r}
dimnames(Titanic)
```

---

* The "names" attribute can be extracted using the function `names()`

* `dimnames()` similarly gives row / column names for matrices and arrays

```{r}
attr(Titanic, "dimnames")
```


---

* For example, column names can be obtained as

```{r}
dimnames(Titanic)[[2]]
```

* There are convenient shortcuts called `rownames()` and `colnames()`

```{r}
colnames(Titanic)
```

---

* In fact, we can easily verify that this is what `colnames()` is doing by printing it

```{r}
colnames
```

* All R functions can be easily inspected in this way

---

* Another very important attribute is "class"

* For example, the return value of `lm()` has class "lm"

```{r}
attr(fm2, "class")
```

* The class of an object can also be obtained using the function `class()`

```{r}
class(fm2)
```

---

* The class of an object can (usually) be "removed" by setting it to
  `NULL` 
  
* This is not something you should actually do!

```{r}
class(fm2) <- NULL
```

* Such objects will no longer have a "class" attribute

```{r}
attr(fm2, "class")
```

* However, it will still have a class (implicitly)

```{r}
class(fm2)
```

<!-- * We will discuss the class attribute in more details next day -->

---

layout: true

# Class, generic functions, and methods

---

* The class of an object is fundamental to how R works

* Every object in R must have a class

* This is true even if the object does not have a class attribute


```{r}
class(colnames)
attr(colnames, "class")
class(Titanic)
attr(Titanic, "class")
```

---

* The main use of the class of an object is in how _generic functions_ behave

* Generic functions are intended to perform general tasks, like

	- `print()`

	- `plot()`

	- `summary()`

* But details of what these functions should do depends on the input


---

```{r}
print(Titanic[, , 1, 1]) 
fm1 <- lm(pm10 ~ no2, aqi, subset = (no2 < 100))
print(fm1)
```

---

```{r}
summary(Titanic) 
```

---

```{r}
summary(fm1)
```

---

* But suppose we make a copy of `fm1` and remove the class attribute from it

```{r}
fm2 <- fm1
class(fm2) <- NULL
class(fm1)
class(fm2)
```

* `fm1` and `fm2` represent the same model fit

* But the different class means that `print()` and `summary()` behave
  differently

---

```{r}
summary(fm2)
```

---

.scrollable400[

```{r}
print(fm2)
```

]

* This kind of customized output is achieved by _methods_

---

* Methods are specific implementations of a generic function customized to its input

* The appropriate method is chosen by looking at the _class_ of the input argument

--

* The methods available for a generic function can be obtained using the `methods()` function

.scrollable400[

```{r}
methods("summary")
```

]

---

.scrollable500[

```{r}
methods("print") # similar but much longer list
```

]

---

* All available methods for a given class can be similarly obtained

```{r}
methods(class = "lm")
```

---

* The name of a specific method appears to have the form `generic.class`

* However, one should always call the generic function, not the method directly

* This is not OK:

```{r,eval=FALSE}
summary.lm(fm1)
```

* Instead, use

```{r,eval=FALSE}
summary(fm1)
```

* In fact, many methods cannot be called directly because they are "hidden"

---

* This is a form of _Object Oriented Programming_ (OOP) in R

* Python also has OOP, but 

	* Methods are usually tied to a class, not a _generic_ function

	* One notable exception is the `__str__()` method, which is used by `print()`

---

layout: false

# Getting help

* R has an extensive collection of functions (even more if we include
  add-on packages)

* It is impossible for anyone to know them all, or remember details

* Fortunately, R also has an excellent help system

--

* Every function and dataset in R (and add-on packages) must be documented

* The documentation can be accessed by `help(name)` or `?name`

* For example: `help(seq)`, `help(summary)`, etc.

* A more general (but limited) search can be performed using `help.search("search-string")`

---

# Getting help

* How the help is shown depends on the _interface_ being used

* RStudio has a separate help tab (which also allows searching)

--

* However, before using the help system, you should know how methods
  are documented

---

# Help for generic functions and methods

* Generic functions and methods are distinct functions

* They often have different help pages

* In fact, many add-on packages define new methods for generics in
  another package

* These are always documented in a separate help page

---

# Help for generic functions and methods

* To get help for the generic function `summary()`, type `help(summary)`

* To get help for the `summary()` method for "matrix" objects, type `help(summary.matrix)`

* To get help for the `summary()` method for "lm" objects, type `help(summary.lm)`

* The first two happen to be the same help page, but the third is different

--

* This is slightly confusing because you are __not__ supposed to call
  `summary.lm()` directly

* More importantly, there may not actually be a `summary()` method for
  all classes 
  
* For example, "list" objects are handled by a _fallback_ method `summary.default()`

--

* The list of available methods are obtained by `methods("summary")`
  as shown earlier
  
* All these should have a corresponding help page

---

# Help for generic functions and methods

- The system we described is called "S3" (short for "S version 3")

- The documentation refers to specific methods implemented using this system as "S3 methods"

- To make things more complicated, there are other systems of defining
  classes and methods
  
- We will skip the details of these for now


---

# Replacement Functions

* R functions are not allowed to modify its arguments

--

* Consider a function that sets negative inputs to 0
$$
f(x) = \begin{cases}
	0 & \text{ if } x < 0 \cr
	x & \text{ otherwise. }
\end{cases}
$$

--

* We will refer to this as the *ReLU* function

---

# The Scalar ReLU Function in R

```{r}
sReLU <- function(u) {
    if (u < 0) u = 0
    u
}
```

???

Now suppose we want to implement this function for scalar input.

This is a very natural implementation.

--

```{r}
x <- -5
sReLU(x)
x
```


???

Let's see what happens when we call it.

If we just call the function with a variable X as input, it returns
the expected value, but the variable X does not change.

---

# The Scalar ReLU Function in R

```{r}
sReLU <- function(u) {
    if (u < 0) u = 0
    u
}
```

```{r}
x <- -5
y <- sReLU(x)
```

--

```{r}
x
y
```

???

This time, when we assign the value of the function to a new variable
Y, X remains unchanged but Y now has the expected value.

This is of course more or less what we would expect.

---

# The Scalar ReLU Function in Python

```{python,cache=FALSE}
def sReLU(u):
    if u < 0:
        u = 0
    return u
```

--

```{python,cache=FALSE}
x = -5
y = sReLU(x)
```

--

```{python,cache=FALSE}
x
y
```

---

# The Vectorized ReLU Function in Python

```{python,cache=FALSE}
def vReLU(u):
    for i in range(len(u)):
        if u[i] < 0:
            u[i] = 0
    return u
```

--

```{python,cache=FALSE}
from numpy import *
x = random.normal(size = 10)
y = x
x
y
```

---

# The Vectorized ReLU Function in Python

```{python,cache=FALSE}
z = vReLU(y)
```

???

When we now call the function with Y as argument and assign the result
to a NEW variable called Z, several interesting things happen.

--

```{python,cache=FALSE}
z
```

???

First of all, Z has the negative values set to 0, as expected.

--

```{python,cache=FALSE}
y
```

???

But Y _also_ has the same values set to 0.

Even though this behaviour is not consistent with the scalar case, it
is not _too_ surprising, because the argument Y WAS actually modified
in the function.

--

```{python,cache=FALSE}
x
```

???

Perhaps more surprisingly, the value of _X_ has also changed, even
though it was not involved in the call at all.

There is actually a very good reason for this behaviour, and depending
on your PRIOR experience with programming, you may find this perfectly
natural.

But this behaviour is often surprising to beginners, and R avoids it
by following what is known as a FUNCTIONAL programming paradigm, which
says that functions should not modify its arguments.

---

# The Vectorized ReLU Function in R

```{r}
vReLU <- function(u) {
    for (i in seq_len(length(u))) {
        if (u[i] < 0)
            u[i] <- 0
    }
    return(u)
}
```

???

Here is the equivalent function in R.

--

```{r}
x <- rnorm(10)
y <- x
```

--

```{r}
x
y
```

???

As before, we first generate a normal random vector X and make a copy named Y.

---

# The Vectorized ReLU Function in R

```{r}
z <- vReLU(y)
z
```

--

```{r}
y
x
```

???

This time, when we run the function with Y as argument and assign the
result to Z, only Z is modified.

X and Y remain unchanged.

---

# Alternative: Direct Assignment

* We can instead use direct assignment to a 'subset'

```{r}
y[y < 0] <- 0
```

--

```{r}
x
y
```

---

# Functional programming

* R generally follows a _functional programming_ paradigm

* Among other things, this says that functions should not modify its arguments

--

* This is a key difference between Python and R

--

* One consequence: R needs an _unusual_ approach when it needs to modify objects

---

layout: true

# Example: Modifying Names of a Data Frame

---

```{r}
d <- data.frame(1, rnorm(5), rexp(5))
names(d)
```

```{r}
d
```

* Default names are not very nice

---

* Want to change the names to `"Constant", "Normal", "Exponential"`

* Possible solution using the `setNames()` function

```{r}
setNames(d, c("Constant", "Normal", "Exponential"))
```

---

* But names of `d` are not modified by this

```{r}
d
```

---

* Best we can hope for

```{r}
d1 <- setNames(d, c("Constant", "Normal", "Exponential"))
d1
```

---

* In fact, the __dplyr__ package has a more convenient version of this approach

```{r}
d2 <- dplyr::rename(d, Constant = X1, Normal = rnorm.5., Exponential = rexp.5.)
d2
```

---

* The classic R Alternative: Replacement Functions

```{r}
names(d) <- c("Constant", "Normal", "Exponential")
d
```


---

layout: false

# Replacement Functions

* Other similar examples:

```r
y[ y < 0 ] <- 0
class(x)  <- NULL
d$Normal[ d$Normal < 0 ] <- 0
attr(x, "name") <- value
```

* Common feature: "complex" expression on the LHS of the assignment

